---
title: "Getting Started with R, RStudio, & R Markdown for ENVH 556:  Part II -- R Markdown formatting "
author: "Lianne Sheppard"
date: "Created for Winter 2021; printed `r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
urlcolor: blue
---

```{r setup, eval=TRUE, include=FALSE}
#-----setup------

# Set knitr options:
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, messages = FALSE, 
                      tidy.opts=list(width.cutoff = 80, blank = TRUE) )

# Set R option: show only 2 digits when displaying
options(digits = 2)

# Clear workspace of all objects and unload all extra (non-base) packages.
# Since knitr creates a fresh environment, this is only here for development.
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
}

# Load packages using pacman (see below for explanation)
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed
# Key principle:  Only load the packages you will need
pacman::p_load(knitr, readr, dplyr, ggplot2, formatR)

```

# R Markdown's text-related tools

There are some things that you might be used to in other text processors (like 
Word) which aren't easy to do in R Markdown.  For example, you'll note that I 
don't indent my paragraphs; that's because it's too much work to do in R 
Markdown.  When formatting assignments for this class, you should not spend 
much time making your text 'pretty', but your answers should be well-organized. 
This section will cover some (easy) things to do in R Markdown to organize and 
manipulate text.

To make a new paragraph in your knitted document, hit 'enter' twice.  If you
only hit 'enter' once, it can be difficult to distinguish paragraphs because
there is no change in line spacing and no paragraph indent.  

You can make font **bold**, or *italicized*.

You can make numbered lists (see the 
Table of Contents), or bulleted lists:

* Note the space between the asterisk and the text.

* This is one of the few cases where R cares about a space.

You can organize your content using headings, which are indicated by `# ` (see
my section headings).  The more pound symbols, the smaller the heading.  We show
these examples without knitting them:

```
# Title (level 1)
## Subtitle (level 2)  
### Sub-subtitle (level 3)  
#### Sub-sub-subtitle (level 4)
```

The easy page breaks in R Markdown use `\newpage`.  You can also separate
content using a string of asterisks or at least three dashes (`---`).  The
string of asterisks creates a dividing line in the output as you see below.

****************

The three dashes give the same result:

---

Finally, you can put comments in your `.Rmd` file that won't be visible in your
output.  For example, this text doesn't appear in the rendered document: (it
appears twice in the `.Rmd` file; the second one is not enclosed in back quotes
so it is hidden.)

`<!-- This is a text comment and won't appear in the output -->`

<!-- This is a text comment and won't appear in the output -->

# Chunks

You can think of each chunk as a miniature R script.  Within a Markdown
document, your code all needs to be in order (you can't print a plot after your
first paragraph if you don't load the data set until paragraph three). You also
can't use *anything* outside the Markdown document; you need to load packages,
load data, etc. all within this Markdown document. This will become an issue
when you try to knit the document, even if it worked while running code straight
from the `.Rmd`.

There are lots of extra things you can do with chunks while running code
straight from the `.Rmd`.  In the upper-right corner of RStudio's text editor
panel, there is a button labeled 'Run'; the drop-down menu from that has useful
chunk commands.

Below is the basic format of a chunk.  You can make new chunks by pressing
'Ctrl' 'Alt' 'i' in Windows, or 'Cmd' 'Option' 'i' in Macs.  You can also just
type out the symbols below (note that those are tick marks, not apostrophes!).
If you read this and are still confused, you might check out [this help
guide](http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html) as well.

It is good practice to name every chunk with a word following `r`, e.g. 
`{r chunk_name_goes_here, and chunk options go after the comma, options separated by commas}`.  


```{r example_chunk}
#-----example chunk-----------

# Code goes here; output appears below

# By setting the seed, we follow the reproducible research practice that my
# random numbers are the same each time.
set.seed(45)

# Generate random numbers from a Normal distribution and save as a vector.
a <- rnorm(mean=0, sd=2, n=20)
mean(a)

```

The code and output above would not be acceptable in a homework write-up.  You
can use chunk options to change how code and output appear in the knitted
document (but chunk options don't affect anything when you are running code in
the `.Rmd` document). Below, I'll give some examples of chunk options; check out
the knitted document to see what each of them do. Chunk options can be combined
by separating them with commas.

## Chunk options

*eval=FALSE*  

The following code won't be run.

```{r code_chunk_that_will_not_run, eval=FALSE, error=TRUE}
#-----code chunk that will not run----

# This code won't even be run!
mean(a)
mean(I can type whatever I want because this is not run.)

```

*echo=FALSE*  

The following code won't be shown, but it will run and the output will be 
displayed.

```{r example_chunk_that_will_not_be_shown, echo=FALSE}
#-----example chunk that will not be shown-----

# This code won't be shown, but it will be run and output will show up.
mean(a)

```

*results="hide"*  

The following output won't be shown, but the code will still run in R.

```{r example_chunk_that_will_not_show_output, results="hide"}
#-----example chunk that will not show output------

# This output won't be shown, but it will still run in R.
mean(a)

```

*Hiding Absolutely Everything: include=FALSE*

To hide absolutely everything, you could use these chunk options:

`echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.show='hide'`

Or, equivalently, just use this one:

`include=FALSE`

For more information about `include=FALSE`, see: 
https://yihui.name/en/2017/11/knitr-include-false/

It is possible to hide errors and warnings in your knitted document.  **Do not**
leave the `message=FALSE` and `warning=FALSE` options turned on while you are
creating your document because they will hide important messages from R.  Just
add these when you are done coding; for example, using `library()` or `p_load()`
often creates some messages which you may wish to hide in your final document.
In general, we *want* to see the messages and warnings from R as we are coding.

The following chunk won't be shown and it creates output that won't be shown.  
However, the code will still run in R.

```{r example_chunk_hides_everything, include=FALSE}
#-----example chunk hides everything----------

# This would be how to do all your coding for which you don't want to show the
# output in your lab report. All this code runs, but nothing shows up in your
# knitted document. Any variable I create here will be 'remembered' for the rest
# of this knitted document.
mean(a)

```

*Accept errors:  error=TRUE*

Without this chunk option, if there is an error in your code the document won't
knit at all.  This chunk option is a temporary band-aid to help you record and
understand the error; **do not** make all your chunks `error=TRUE` as a way of
bypassing code issues.

```{r example_chunk_accept_errors, error=TRUE}
#-----example chunk accept errors-----

# Sometimes in a work-in-progress, you may want a knitted copy of everything -
# even things that go wrong. This wouldn't be a chunk option to use in a
# polished homework assignment. This is an error because I haven't loaded the
# titanic data in this document yet (regardless of whether it's loaded in my
# RStudio right now)
head(titanic) 
mean(a)

```

## Naming chunks

It is good practice to name your chunks as we have done in the examples above.
Further, if you put a comment ending with at least 4 dashes at the end of the
first comment line, this shows up in the outline section of your document in
RStudio AND makes it easier to read your code in the appendix.  A comment
formatted this way looks like this: `#-----example chunk name-----`

## Plots

To print a plot, you don't need any special chunk options.  

```{r example_plot}
#-----example plot-----

hist(a)

```

However, to hide a plot you need to use `include=FALSE`, which hides all code 
and output (using `results="hide"` won't work on plots).

*include=FALSE to hide a plot and its code*

The following chunk is hidden, as is its output.

```{r example_plot_is_hidden,include=FALSE}
#-----example plot is hidden----------

# Still run in R, though!
hist(a)

```

# In-line code

We saw above that output from chunks is usually not 'pretty' enough for refined
reports.  To include code output in a sentence, we use this format of one
back-tick followed by "r ", a space, and then some R code, and ending with a
second back-tick.   So we can write: The mean of the data is `r mean(a)`.
Ideally you pay attention to scientific reporting standards, for instance by
rounding it to an appropriate number of significant figures: The mean is `r
round(mean(a), 2)`.  This is much better than typing "The mean is 0.217" because
it follows several reproducibility principles.  (See Part I for details.)

In general, only use the in-line code format for printing out one or two
numbers.  If you need a table or plot, that needs to come from a chunk.

# Tables

R Markdown has the ability to create nice tables (nice enough to use for
homework).  There are several different packages for creating tables; these
include `kable`, `stargazer`, and `xtable`. `kable` is the simplest to use and
we show examples in Lab 1. If you wish to develop more sophisticated tables,
explore these other packages.

# Taking advantage of R Markdown's capabilities within the RStudio environment

Here are a few tips and tricks that can help you work quickly in R Markdown, 
which you can't use in a regular R script.

  * Run the current chunk: Ctrl+Shift+Enter (Windows) or Command+Option+c (Mac) 
or press the green 'play button' in the upper-right hand corner of the chunk
  * Run the next chunk: Ctrl+Alt+n (Windows) or Command+Option+n (Mac)
  * Run all chunks above: Ctrl+Alt+p (Windows) or press the upside-down triangle 
in the upper-right hand corner of the chunk
  * Run all chunks: Ctrl+Alt+r (Windows) or go through the 'Run' button drop-down 
(see below)

There are lots more running options if you select the drop-down of the 'Run'
button on the right side of the toolbar for RStudio's text editor panel.  You
can use the arrows to the left of the 'Run' button to help navigate your
document, especially if you have a lot of text.

RStudio's text editor also has a primitive spell-checker which is good at
checking only regular text and not any code or special symbols.  To use it,
press the button to the right of the save button in the toolbar for the text
editor panel.  It has 'ABC' and a green check mark.  If you want something more
sophisticated which will also check grammar, I would recommend knitting to a
Word document and just using the Word spell-checker. Just don't paste code from
Word back into your text editor, as MS Word may modify characters such as
quotes, which will likely break your code.

## Tips on knitting to a PDF

You may want to knit your R Markdown directly to a `.pdf` to turn in.  This is
very easy; although sometimes you may need to install an extra package.  To
check whether your computer is already set up to knit to `.pdf`, try selecting the
drop-down from the 'Knit' button and choosing 'Knit to PDF' (or change your
header, as described below).  If that doesn't work, you'll have to install the
`tinytex` package. You can do this within R if you use the
[tinytex](https://yihui.name/tinytex/) package, which is a minimal \TeX package
for R users.

```
install.packages('tinytex', repos = 'http://cran.r-project.org')
tinytex::install_tinytex()
```

You would only need to run these commands once. You would not normally include 
them in a regular R script, as doing so would needlessly repeat the 
installation, unless you take special precautions, such as wrapping the code 
in an `if` block as as shown below.

```{r install_tinytex, eval=FALSE, echo=TRUE}
#-----install tinytex-----

if (!file.exists(Sys.which('texi2dvi'))) {
    # This can take a long time to run
    install.packages('tinytex', repos = 'http://cran.r-project.org')
    tinytex::install_tinytex(force=TRUE)
    if (tinytex:::is_tinytex() != TRUE) warning('tinytex failed to install!')
}

```

If you need a more complete \TeX package for other uses, you will need to 
download and install more software, separately from R. For Windows users, 
there is an awesome tutorial 
[here](https://medium.com/@sorenlind/create-pdf-reports-using-r-r-markdown-latex-and-knitr-on-windows-10-952b0c48bfa9), 
and there is also a [Mac version](https://medium.com/@sorenlind/create-pdf-reports-using-r-r-markdown-latex-and-knitr-on-macos-high-sierra-e7b5705c9fd).  
This will walk you through what you need to install (it also goes through 
installing R and RStudio, but you can just skip that).  Once everything is 
installed, you can set your output to `pdf_document` instead of `html_document` 
to switch between `.html` and `.pdf` files.

## Using \LaTeX

If you want to make your text clearer by using nicely formatted equations, you 
can use the math language \LaTeX to write Greek letters and equations.  Below 
are some examples that you can copy and paste, or just Google it.  Be prepared 
to have a little patience, because \LaTeX is pretty picky.  This is *not* an 
expectation for ENVH 556, it's just something extra for those of you who are 
interested.  This is also an excuse to plug a 
[useful website for finding \LaTeX commands](http://detexify.kirelabs.org/classify.html).

$\alpha$

$\alpha_0$

$\alpha_1$

$\beta$

$\pi$

$x_{Exposure}$

$e^{\beta}$

So we can write out models:

$\log(x_{Disease})=\alpha_0+\beta x_{Exposure}$ 

or log($x_{Disease}$) = $\alpha_0$ + $\beta$ $x_{Exposure}$

# Package Namespace Conflicts

Sometimes we need to load a package which will have function names which 
conflict with packages we have already loaded. When you call that function, 
R might run the function from one package when you intended it to be run from 
another. You may get strange results or errors because of this. They can be
difficult to solve because the true source of the conflict is not clear.

For example, let's say you load the `plyr` package and then load the `dplyr` 
package, as shown below.

```{r load_plyr_and_dplyr, eval=TRUE, echo=TRUE}
#-----load plyr and dplyr-----

pacman::p_load(plyr, dplyr)

```

R may even warn you about the conflicts as the second package is loaded:

```
Attaching package: ‘dplyr’

The following objects are masked from ‘package:plyr’:

    arrange, count, desc, failwith, id, mutate, rename, summarise, summarize

The following objects are masked from ‘package:stats’:

    filter, lag

The following objects are masked from ‘package:base’:

    intersect, setdiff, setequal, union
```

If you want to use the function `count`, you may not get the one you intend, 
as both of these packages contain `count`, but the function arguments are
different. This could give you an error if your argument list did not match
what was expected by the function. This situation is called a namespace conflict.
R includes tools to examine these conflicts.

```{r examine_namespace_conflict, eval=TRUE, echo=TRUE}
#-----examine namespace conflict-----

# Find namespaces which contain "count"
getAnywhere(count)$where

# Find the expected arguments for plyr::count
args(plyr::count)

# Find the expected arguments for dplyr::count
args(dplyr::count)

```

We can avoid this conflict by only loading one package that contains the
`count` function, or by specifying the package when we call the function.

```{r calling_functions_by_package_name, echo=TRUE, eval=TRUE}
#-----calling functions by package name-----

# Use plyr's count function and dplyr's count function to count the number of
# observations by species in the iris dataset

# Make sure the dataset is loaded
data(iris)

# plyr's count function expects the "vars" grouping argument (argument #2) as a 
# quoted character vector of variable names
plyr::count(iris, "Species")

# If we use dplyr's count function in the same way, we get strange results
dplyr::count(iris, "Species")

# dplyr's count function expects the (unnamed) grouping argument as an unquoted
# variable name or a vector of data values
dplyr::count(iris, Species)

```

## A Note about Generations and Families of Packages

We have used the `plyr` and `dplyr` packages above and have seen there was a
conflict when we loaded both of them at once. As it turns out, `dplyr` is
a package which was developed to take the place of `plyr`. So, why would we load
both packages? One reason is that we may be accustomed to using the older
package but also need new features of the later package. Another reason might 
be that the newer package left out a feature we needed from the old one.

If this situation sounds a little messy to you, it should. As you can imagine,
we start to get into trouble when we use packages from different generations or
families. If `dplyr` was meant to make `plyr` obsolete, then we should switch to
`dplyr` and update our code to match `dplyr`'s syntax. `dplyr` is in the family
of packages called ["The Tidyverse"](https://www.tidyverse.org/) since they are developed by a common group 
of developers and also since they share a common philosophy and vision.

You can see the list of packages in the `tidyverse` family with the following
command:

```{r show_tidyverse_functions, echo=TRUE, eval=TRUE}
#-----show tidyverse functions-----

pacman::p_load(tidyverse)
tidyverse::tidyverse_packages()

```

You will see `dplyr` in the list but not `plyr`. Similarly, `tidyr` has taken
the place of `reshape2`. Fortunately in this case the functions in 
`tidyr` have new names. Only a dataset is common to both packages. It contains
the same data values, as you can confirm with the following code:

```{r compare_reshape2_and_tidyr, echo=TRUE, eval=TRUE}
#-----compare reshape2 and tidyr-----

# Find the differences in the reshape2 and tidyr packages

# Load packages
pacman::p_load(reshape2, tidyr)

# Show the functions in each package (commented out for brevity)
#lsf.str("package:reshape2")
#lsf.str("package:tidyr")

# Find the objects which are in tidyr but not in reshape2
ls("package:tidyr")[ls("package:tidyr") %in% ls("package:reshape2")]

# Find the class of the "smiths" object in each package
class(reshape2::smiths)
class(tidyr::smiths)

# See if the two versions of the "smith" dataset are identical
identical(reshape2::smiths, tidyr::smiths)

# See if the two versions of the "smith" dataset are otherwise equivalent
suppressWarnings(all_equal(reshape2::smiths, tidyr::smiths, convert = TRUE))

```


Sometimes, the packages we use will load other packages as "dependencies". We
can't easily control which packages are loaded indirectly like this. Sometimes
we end up loading packages which create namespace conflicts unknowingly. So, if
you see strange errors when calling a function, especially those which refer to
misuse of function arguments, think about examining potential namespace
conflicts and explicitly calling the function from the intended package.

Four best practices to apply are:

  * Try to standardize on common families of packages when possible.
  * Use the latest generation of packages from your chosen family of packages.
  * Only load the minimum number of packages into memory that you will need.
  * Explicitly use package names when calling functions if you must in order to 
  avoid namespace conflicts

# Additional Resources

* For more background on why and how to use R Markdown to create reproducible 
scientific papers, see 
[Michael Frank's YouTube video](https://www.youtube.com/watch?v=Nj9J5iCSMB0), 
or the [html document it reviews](https://cdn.rawgit.com/mcfrank/openscience_tutorial/fec71433/rmarkdown_handout.html), 
and the [R Markdown (.Rmd) document](https://github.com/mcfrank/openscience_tutorial/blob/master/rmarkdown_handout.Rmd) 
that created it.
* Try to find time to read more of the [R for Data
Science](https://r4ds.had.co.nz/) book by Hadley Wickham and Garrett Grolemund.
In particular, the "Explore" section (chapters 2-8) gives an excellent overview
of tools we will use in ENVH 556.

# Appendix

```{r session_info}
#-----session_info----
# Print the session information
sessionInfo()

```


```{r code_appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, include=T}
#-----code appendix------------
```

