---
title: "Lab Report Guidelines & Template"
subtitle: "Created for Winter 2021"
author: "ENVH 556 Instructors"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    html_document:
        fig_caption: yes
        toc: true
        toc_depth: 3
        number_sections: true

---

```{r setup, include=FALSE}
#-----setup-----

# Some useful knitr options:
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, tidy = FALSE, 
  tidy.opts=list(width.cutoff=80, blank = FALSE), cache = FALSE, messages = FALSE)

# Set R option:  here we show only 2 digits when displaying
#options(digits = 2)

# Clear workspace of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
}

# Load key packages using pacman (see below for explanation)
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed
# Key principle:  Only load the packages you will need
pacman::p_load(knitr)

```

# Lab Report Overview, Guidelines & Suggestions {-}
<!--consider adding a section to cover "higher standards for peer reviewed journal articles" -->

Each lab report should be written using R Markdown following the outline in this document and adhering to reproducible research principles. Although you will include a code Appendix that will allow an independent replication of your results, all material that you want to be considered in grading should be in the main part of your report. 

Overall your lab report should be concise, but long enough that it stands alone as a comprehensible document to the reader. We have given suggestions below about the length of each section.

Remember to define variables/acronyms/abbreviations (e.g. "nitrogen oxide (NO)" or "nitrogen dioxide (NO2)") once, on first use, and specify units everywhere appropriate (i.e. in text and in figures/tables).

You can use this document as a template for your reports. Just delete this "Lab Report Overview, Guidelines & Suggestions" section (and sub-sections) and replace the text throughout the rest of the document.

## Formatting {-}
There are many challenging aspects of R Markdown formatting. Generally, these report instructions are intended to reduce the burden of making "publication-quality" figures, tables, and text. While we are allowing for aesthetic compromises, we are not suggesting you take any shortcuts in the material or content of your report.The following examples should help illustrate the balance we are trying to acheive: 

  * Tables do not need to look "pretty" and you are not required to follow specific instructions for spacing, borders, indentation, etc. However, your tables still need to be thoughtfully presented, for example: the variables you choose to display, whether the table is oriented by row or column, if the layout promotes the comparisons of interest (e.g. column order), and the clarity of the column/row names. Ideally the values you report should have the appropriate number of significant digits, but because this can be challenging to implement in R Markdown, we won't penalize you for reporting too many significant digits.
  
  * Figure axis labels do not need to contain special characters. However, the labels need to clearly convey what you are trying to communicate. For example, you are not required to present units such as "PM$_{2.5}\space(\mu g/m^3)$" in axis labels, but you could instead write "PM2.5 (ug/m3)" or "PM2.5 (ug/m^3)" or "PM2.5 (ug*m^-3)." 

You may compile your R Markdown file into a word, pdf, or html file, but we suggest `.html` files as they are the easiest to format and render. The [HTML Chapter](https://bookdown.org/yihui/rmarkdown/html-document.html) of ["R Markdown: The Definitive Guide"](https://bookdown.org/yihui/rmarkdown/) is a helpful resource.

## Drafting your report {-}
Writing drafts of the introduction, data, and methods sections before the data analysis begins can be very useful in guiding your analysis and *save time*. A surprisingly large amount of the report can be written before the detailed analysis has begun. While drafting your lab report you can render a `.Rmd` with the `eval = FALSE` option to prevent the code from being run before you are ready.

## Some ideas for sanity checks as you work {-}

* How do you know you are analyzing the data you think you are analyzing?  
     + Have you verified the sample size is correct at each data massaging step?
     + How do you know data have been merged correctly?

* Make sure you understand the data:  
    + What low-dimensional summaries of the data have you tabulated/plotted?  Do they make sense given your understanding of what you *should* have?
    + What do you expect to see if the data are purely noise?

## Guidance on writing and formatting your code {-}

1. In general you should finalize your code like you finalize your scientific writing:  It should be clear, concise, and nicely formatted. Incorporate the following **formatting strategies** into your code in your assignments:

    A. *Be clear about what each chunk does.* You can do this by 1) giving chunks descriptive names/header comments and 2) commenting the body of a chunk.  See examples in the course R Markdown files.
    
    B. *Use descriptive names for dataframes and variables*, including referring to variables by name rather than position when indexing.
    
    C. *Write functions* for blocks of code that are reused extensively, using descriptive function names. A good rule of thumb is not to copy/paste bits of code more than 3 times. 
    
    D. *Show dataframes, where appropriate.* If a block of code processes data, it may be helpful to print a few lines of data before and after to show how the processing worked. (e.g., Use `head()` and/or `tail()` or `tidyverse` options).
    
    E. *Make sure you don't have run-on lines* in your compiled code. Check for this and fix it before finalizing your assignment. Mostly you can rely on tools to automatically reformat the lines because these tools will accomplish what you want (though sometimes they don't). R Markdown has formatting options (e.g. `tidy=TRUE, tidy.opts=list(width.cutoff=80)` ) that help control the width of code rendered in reports. Another strategy to automatically reformat code, is to highlight the lines you want to reformat and then:
      i. For lines of code use: Cmd+Shift+A for macs, or Cntrl+Shift+A for windows.
      ii. For comments use: Cmd+Shift+/ or Cntrl+Shift+/.  
      
    F. *Delete extraneous code and comments* unless you have a reason to keep them in your R Markdown file, for instance to remind you of something important. In the lab R Markdown files we have been intentionally leaving in some commented sections to show you alternative ways to do things, help you understand what we've done, or sometimes for other reasons. When you copy that code for your own purposes, delete the extra material you don't need. Your comments should be informative and orienting, but not excessive.
    
    G. *Break up your code and don't make code chunks too long.* To  make your code more readable, try not to include chunks of code that are longer than 1/3 of a page.
      i. Use header comment lines at the beginning of code chunks as signposts, e.g. `#----this is a header line----` You can decide your personal style for this. Also if you put at least 4 dashes at the end of the header line, RStudio incorporates this line into your outline. 
      ii. Use line spacing and indentation.
      iii. Insert a blank line before each comment line that describes what the next section of code does.   
    
    H. *Adopt a consistent style* and work to continuously improve your style. Here are three popular and suggested style guides: 
      i. The [Google style guide for R
code](https://google.github.io/styleguide/Rguide.xml), 
      ii. The [Advanced R Guide (Hadley Wickam's tweaks of the Google style guide)](http://adv-r.had.co.nz/Style.html)
      iii. The [tidyverse style guide](https://style.tidyverse.org/).  
      Your style will improve with practice, so make it a goal to progressively improve with each assignment. There are many details to learn!

2. Some of the demonstrations this course provided in the lab assignments are for your learning, but don't need to be brought forward into your lab write-up. For instance, once you know how to write a function and are convinced that your function works correctly, there is no need to use the non-function version. As an example, Lab 4 demonstrates cross-validation (CV) manually in order to convey thoroughly cross-validation ideas. Then the lab shows how to write functions and turn the CV code into a function. The lab also follows the good practice of checking the function by demonstrating that the function results are identical to doing it manually. Once we are convinced that the function works properly, there is no need to continue to use the manual cross-validation since it is always better to use a function for a repetitive task.

3. Pay attention to the overall organization of your analyses and code. For instance, if you find you need a new package, add your command to load it at the beginning of the R Markdown file in the chunk where you load all the other packages. As another example, be intentional about the choice to repeat some aspect of your code vs. just work with a saved object. Similarly, once you know how to write functions, use them instead of repeating code.

4. It can be useful to keep in your R Markdown file a record of the exploratory work you did but that doesn't show up in your final report. It is good practice with any new dataset and/or data analysis to make sure you really understand your data before you launch into your full analysis. Often you won't put very much or any of this exploratory work in a final write-up. However, it can be very helpful for you to keep a record of the kinds of descriptive analyses you did by leaving unevaluated code in your R Markdown file. This is one way that you can keep track of what you looked at in the past. Also it is a way that those following your work can check that you are making sure you are understanding key features of your data and analyses. 


## Lab Report Outline {-}
Your report should include the following sections:

1. Introduction
    * Background and rationale
    * Objectives
2. Data & Methods
    * Data description
    * Statistical analysis approach
3. Results
4. Discussion and Conclusions
5. References  
6. Appendices
    * Extended data analyses (optional)
    * Session information, code, function definitions (required) 

Instructions for each section are included below.  We provide a suggested length for each section as a general guide.  Overall your lab report should be as short as possible while still meeting the objectives of the lab.

---


# Introduction
**Suggested length: 1-2 paragraphs**

Use this section to provide an introductory paragraph (or two) summarizing the purpose of the lab and a short scientific description of the data. The introduction provides background and context for you analysis. Clearly state the underlying scientific question(s) you hope to address. A literature review is not necessary in your lab report. Potential topics in this section include:

* Background and rationale:
    + Why are we doing this? (i.e., what is the purpose of the lab?)
    + What data are we using and what do they represent scientifically?  

* Objectives
    + What are we trying to learn?
    + What outcomes would constitute success?

---

# Data & Methods
**Suggested length: ~1-3 paragraphs**.  
This section should be only as long as needed to convey the key aspects of your data and analysis approach that are needed to understand the results and interpretation.   

## Data description 
This section covers all aspects of data description that don't fit naturally into the results section and weren't already mentioned in the introduction. It can be brief or omitted in your lab write-up if the data and context are already clear from your introduction.  As needed, use this section to provide any additional background and context required, for example, related to the dataset you are using. This may include a description of the experiments or observational study design with mention of how these are expected to answer the question (_a priori_ expectations about the results).

## Statistical analysis approach 
This section tells the reader the approach you took to answer the questions posed. Make sure to include a description of your analysis approach(es) with sufficient details to put the results in context.  To help keep your lab write-up brief, you may assume that the reader understands concepts and tools developed in class.  However, you may need to provide some details of your implementation.  For instance, you might state that you did cross-validation without explaining the details.  However, you should note whether it was leave-one-out or k-fold, where you indicate the value for k.

## Additional questions to consider {-}  
The following questions cover topics that often need to be documented in an analysis write-up and may be particularly useful for you in future analysis documents.  They won't necessarily apply to every lab report and answers to some of them fit more naturally at the beginning of your results section.    

* Data description  
  + Where did our data come from? (Supply URLs as appropriate)
  + How many samples/subjects?
  + How many measurements per subject/sample? 
  + What covariates are important?

* Data processing
  + If the data will be filtered, stratified/grouped, what cutoffs will be used?
  + Were there any zero concentrations or concentration values less than the limit of detection (<LOD) that you dropped, imputed, estimated, etc.?

* Analytical approach & methods
  + What statistical tests are being used?
  + What sensitivity analyses are planned?

---

# Results
**Suggested length: No more than 1 paragraph per distinct topic**, e.g. per figure or table 

**Maximum Tables & Figures: 5 total**

Summarize the results of the analysis with text, tables and figures. Use the results (and discussion) sections to answer all the questions posed in the assignment. The results section should provide an *objective summary* of the findings; interpretation goes in the discussion section. Anticipate questions from readers and answer them in your presentation of your findings. While signposting the questions is helpful for grading and therefore appreciated, please organize this section as readable text rather than Q&A responses to the questions in the lab.

In most lab assignments you should not need more than 2-5 figures or tables.Your lab write-up is limited to a maximum of 5 figures and/or tables in the main report. Generally, these figures and tables will be presented in the results, but certain circumstances may warrant they appear elsewhere - use your judgment. Additional figures and tables can be provided in the appendix if desired, but an important learning objective for this course is thinking about what figures/tables are presented in the main body, the  appendix/supplemental information, or not at all. Be thoughtful about how you design your tables and figures to convey the most important information. Cite tables and figures as appropriate to provide support for your written answers to the questions asked in the lab.

Here are some strategies to maximize the effectiveness of your tables/figures:

  *	Figures and tables should be strategically selected to present your key analyses and findings
  *	All figures and tables need appropriately descriptive captions/titles. 
  * Pay attention to table layout, column order, labels, headings for rows/columns, and ensure that all necessary information is included (e.g. in footnotes). Omit redundant columns as well as those with no relevant information or zero variability. Additional resources for creating HTML tables include: 
    + kableExtra [vingette](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) 
    + R Markdown Cookbook [Chapter 10: Tables](https://bookdown.org/yihui/rmarkdown-cookbook/kable.html)

  *	Structure tables to facilitate the comparisons of interest. Note that it is easier to compare numbers that are close to each other and to make comparisons of numbers in columns. 
  * Think about what is inferred from each plot you show:
    + Describe what allows you to make this inference.
    + Think about what the plot would look like if there were no structure in the data.

---

# Discussion and Conclusions
**Suggested length: Up to 1 paragraph per objective with possibly 1-2 additional paragraphs to address other topics such as limitations, future work, or conclusions** 

*In general*, the discussion section provides the context for the results. This is the place for interpretation in light of your objectives. You can also include a discussion of the implications of the findings, bring in additional perspectives, and/or ask questions you have. For instance, this is a place to discuss the limitations of the data and your analysis. You could compare your results to other studies. You may also wish to address data quality issues, exceptions, outliers, caveats or limitations of the approach, and possible next steps.

*For your lab report*, it is useful to organize the discussion around the objectives. For each objective, include a statement or two about how that objective has or hasn't been met. Keep your reporting of the objectives in the same order as your presentation in the results section. Describe the conclusions to be drawn from the results.  As appropriate, you may also wish to address other topics described in the previous paragraph.

---

# References

Include any references beyond those assigned in the course or lab. This section may be omitted if you don't use references beyond those in the assignment.

While not required for class, there are tools to integrate reference manager libraries (e.g. [Zotero](https://www.zotero.org/)) into `.Rmd` files. RStudio's instructions for [bibliographies and citations](https://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html) in R Markdown will get you started.

---

# Appendices

## Appendix 1 (optional): Supplemental results

Include any additional details that don't belong in the primary results but need to be included. This section can cover: 

  * Steps for reading and preprocessing the data
  * Descriptive data analyses and verification that you are analyzing the data you intend to analyze
  * Additional descriptive analyses
  * Additional inferential analyses
  * Additional figures and tables  

## Appendix 2 (required): Session information, embedded code, and functions

Include the session information, embedded code, and functions defined in the `.Rmd` in a code appendix. This appendix output is also shown in the compiled html file for these guidelines. Note that there is very little code embedded in these guidelines and no functions, so the second two chunks produce little or no oputput. Generally your lab reports will have much longer code appendices.

### Session information

```{r session info}
#-----session info------
sessionInfo()

```

### Code embedded in the R Markdown file

```{r code appendix, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60), include=T}
#-----code appendix-------

```

### Functions defined in the R Markdown file

```{r functions.defined.in.this.Rmd, eval = TRUE}
#-----functions defined in this Rmd ------------

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lapply(c(lsf.str()), getAnywhere)

```

---

