---
title: "Week5Lecture: Variance Components"
author: "Lianne Sheppard for ENVH 556"
date: "Winter 2019; Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        fig_caption: yes
        toc: true
        toc_depth: 3
        number_sections: true
        
---

<!--Basic document set-up goes here  -->

```{r setup, include=FALSE}
#-------------r.setup-------------
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear.workspace, eval=FALSE, echo=FALSE}
#---------clear.workspace------------
# Clear the environment without clearing knitr
#
# This chunk is useful for code development because it simulates the knitr
# environment. Run it as a code chunk when testing. When knitr is run, it uses a
# fresh, clean environment, so we set eval=FALSE to disable this chunk when
# rendering.

# Clear workspace of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}

```


```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#----------------load.libraries.pacman----
# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.  Some reasons for packages:
# knitr:  kable()
# ggplot2: part of tidyverse
# readr: part of tidyverse
# dplyr: part of tidyverse
# multcomp:  glht
# modelr:  part of tidyverse and need for add_predictions and add_residuals
# boot:  cv tools are available
# Hmisc:  describe
# lme4: mixed models and random effects ANOVA
# VCA for variance component analysis (note it masks several lme4 objects:  
#      fixef, getL, ranef)
pacman::p_load(tidyverse, knitr, lme4, VCA)  
```

```{r read.data.from.a.web.site, eval=TRUE, echo=FALSE, include = FALSE}
#-----read.data.from.a.web.site--------
# Download the data file from a web site if it is not already downloaded, then
# read in the file

datapath <- "Datasets"
dir.create(datapath, showWarnings=FALSE, recursive = TRUE)

snapshot.file <- "allseasonsR.rds"
snapshot.path <- file.path(datapath, snapshot.file)

# Only download the file if it's not already present
if (!file.exists(snapshot.path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 snapshot.file, sep = '/')
    download.file(url = url, destfile = snapshot.path)
}

# Output a warning message if the file cannot be found
if (file.exists(snapshot.path)) {
    snapshot <- readRDS(file = snapshot.path)
} else warning(paste("Can't find", snapshot.file, "!"))

```

```{r read.data.from.a.web.siteDEMS, eval=TRUE, echo = FALSE, include = FALSE}
#-----read.data.from.a.web.siteDEMS--------
# Download the data file from a web site if it is not already downloaded, then
# read in the file

DEMS.file <- "DEMSCombinedPersonal.rds"
DEMS.path <- file.path(datapath, DEMS.file)

# Only download the file if it's not already present
if (!file.exists(DEMS.path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 DEMS.file, sep = '/')
    download.file(url = url, destfile = DEMS.path)
}

# Output a warning message if the file cannot be found
if (file.exists(DEMS.path)) {
    DEMS <- readRDS(file = DEMS.path)
} else warning(paste("Can't find", DEMS.file, "!"))

```

# R commands & output for Variance Components Lecture Slides

```{r DEMS data setup}
# DEMS data setup ------
# create the varaiable ln_ecdata among non-missing
DEMS <- DEMS %>% filter(ecdata > 0) %>%
    mutate(ln_ecdata = log(ecdata)) 
    
```

## slide 10:  Plot the DEMS facility example 


```{r dotplot of DEMS}
# dotplot of DEMS ----------
ggplot(DEMS, aes(x = facilityno, y = ln_ecdata)) +
    geom_boxplot(aes(x=facilityno, group = facilityno),
                 outlier.color = NA, 
                 width = .3) +
    geom_dotplot(aes(group = facilityno), binaxis = "y", binwidth = .05, 
                 stackdir =  "center", color = "red", fill = NA, dotsize = 2) +
    labs(title = "Boxplots with Overlaid Dotplots\nfor Ln(REC) by Facility Number",
         y = "ln(REC) in ln(ug/m3)",
         x = "Facility ID")
```
## DEMS Stratified by facility ID (not a slide)

```{r DEMS tidyverse stratified summary}
# DEMSvtidyverse stratified summary-----
vc_summary <- DEMS %>%
    group_by(facilityid) %>%
    summarize( N = n(), 
               mean = mean(ln_ecdata),
               sd = sd(ln_ecdata)) 
# kludge to get the total
vc_total <- DEMS %>%
    summarize( N = n(), 
               mean = mean(ln_ecdata),
               sd = sd(ln_ecdata))
vc_total <- cbind(facilityid = "Total", vc_total)

kable(rbind(vc_summary,vc_total), digits = 2)
```

## slide 12:  Plot the example of traffic gradient site as one grouping variable in the snapshot data

```{r Snapshot data setup}
# Snapshot data setup -----------
# get the gradient sites
gradient_sites <- snapshot %>% 
    filter(season == 2)  %>% 
    group_by(group_loc) %>%
    summarize(count = n()) %>%
    filter( count > 3)
# fall dataset for variance components
# keeps only gradient sites with at least 4 observations
fvc <- snapshot %>% filter(season == 2, group_loc %in% gradient_sites$group_loc) 
```

```{r dotplot of fvc}
# dotplot of fvc ----------
ggplot(fvc, aes(x = group_loc, y = ln_nox)) +
    geom_boxplot(aes(x=group_loc, group = group_loc),
                 outlier.color = NA, 
                 width = .6) +
    geom_dotplot(aes(group = group_loc), binaxis = "y", binwidth = .01, s
                 tackdir =  "center", color = "red", fill = NA, dotsize = 2) +
    labs(title = "Boxplots with Overlaid Dotplots\nfor Ln(NOx) in Fall by Gradient Site Index",
         y = "ln(NOx) in ln(ppb)",
         x = "Gradient Site Index")
```
## Slide 18:  Stratified by gradient location

```{r tidyverse stratified summary snapshot}
# tidyverse stratified summary-----
vc_summary <- fvc %>%
    group_by(group_loc) %>%
    summarize( N = n(), 
               mean = mean(ln_nox),
               sd = sd(ln_nox)) 
#vc_summary <- recode_factor(group_loc,.default)
vc_total <- fvc %>%
    summarize( N = n(), 
               mean = mean(ln_nox),
               sd = sd(ln_nox))
vc_total <- cbind(group_loc = 100, vc_total)

kable(rbind(vc_summary,vc_total), digits = 3)
```

## Slide 19:  Assessing between and within site variability

```{r gradient variability}
# slide 19: gradient variability
vc_var <- vc_summary %>%
    summarize(N_sites = n(),
              median_grand = median(mean),
              mean_grand = mean(mean),
              median_sd = median(sd), 
              within_sd = sqrt(mean(sd^2)),
              between_sd = sd(mean)
              )
vc_var

paste("Between-group variance estimate(SD scale) = ", round(vc_var$between_sd,3))
paste("Within-group  variance estimate(SD scale) = ", round(vc_var$within_sd,3))
paste("Total         variance estimate(SD scale) = ", round(sqrt(vc_var$within_sd^2 + 
                                                                    vc_var$between_sd^2),3))
```

## Dropped from lecture:  Site as a fixed effect

Needs development to focus this result in comparison to the random effect
result.  Note that this is a regression model based approach to the previous
more descriptive approach.

```{r}
# slide 16: site as a fixed effect
fixed_ef <- lm(ln_nox ~ as.factor(group_loc), data = fvc)
summary(fixed_ef)
```

### Dropped from lecture:  Same in ANOVA (fixed effect model)


```{r}
#slide 17:  same in ANOVA
anova(fixed_ef)
```

## Slide 21 + 22:  Variance components with VCA

VCA allows the MOM estimates and easy comparison with REML estimates of variance
components.

```{r}
#slide 21+22:  VCA immplementation
# visualize data
varPlot(form = ln_nox~group_loc, Data=fvc)
## different ways to estimate variance components
fit_MOM <- anovaVCA(ln_nox ~ group_loc, Data = fvc)
fit_MOM
fit_REML <- fitVCA(ln_nox ~ group_loc, Data = fvc, method = "REML")
fit_REML
# the following gives the same result so is not repeated
#fit_ANOVA <- fitVCA(ln_nox ~ group_loc, Data = fvc, method = "anova")
#fit_ANOVA
```


## Slide 23:  Estimation using lme

A key point here that the `lmer` REML estimates are the same as from `VCA`.
Discuss why we use ML at all -- basicially to easily compare models.

```{r REML snapshot}
# REML with snapshot -----------
lmefit_REML <- lmer(ln_nox ~ (1|group_loc), data = fvc, REML = TRUE)
summary(lmefit_REML)
```

```{r ML snapshot}
# ML with snapshot ----------------
lmefit_ML <- lmer(ln_nox ~ (1|group_loc), data = fvc, REML = FALSE)
summary(lmefit_ML)
```


## Dropped from lecture:  Same examples with DEMS 

### DEMS with VCA


```{r}
#slide 26+27:  VCA immplementation in DEMS
# visualize data
varPlot(form = ln_ecdata~facilityid, Data=DEMS)
## different ways to estimate variance components
fit_MOM <- anovaVCA(ln_ecdata ~ facilityid, Data = DEMS)
fit_MOM
fit_REML <- fitVCA(ln_ecdata ~ facilityid, Data = DEMS, method = "REML")
fit_REML
fit_ANOVA <- fitVCA(ln_ecdata ~ facilityid, Data = DEMS, method = "anova")
fit_ANOVA
```

## Slide 26:  DEMS comparison of estimates

I did not develop the R coding for this.  We would need to use the code for the
snapshot data, now applied to DEMS, in this section

## Slide 30:  Mixed model with the Snapshot data: Pop_5000

The following analyses address how the two variance components change when
different covariates are in the model.  As was done in the Peretz paper, think
about whether we expect an impact on the variance components and whether the
impact should affect the between site or within site variance estimates.

Note:  Because one might want to do some model selection comparisons with these
analyses, I used the ML option (i.e. REML = FALSE).

```{r mixed model with Pop_5000}
# mixed model with Pop_5000 ----------
mixedfit1_ML <- lmer(ln_nox ~ Pop_5000 + (1|group_loc), data = fvc, REML = FALSE)
summary(mixedfit1_ML)
```

## Slide 31:  Mixed model with the Snapshot data: D2R

```{r mixed model with D2R}
# mixed model with D2R ----------
mixedfit2_ML <- lmer(ln_nox ~ D2R + (1|group_loc), data = fvc, REML = FALSE)
summary(mixedfit2_ML)
```


## Dropped from lecture:  Mixed model with the Snapshot data: two covariates

Note: to compare models you need to use ML because of how the variance
components are estimated to be unbiased in REML.  For final estimates we should
report REML.

```{r}
mixedfit3_ML <- lmer(ln_nox ~ Pop_5000 + D2Comm + (1|group_loc), data = fvc, REML = FALSE)
summary(mixedfit3_ML)
```


## Slide 35:  DEMS V Table 3 Facility A example U1 groups

Verifying:  This result has the same number of observations (108) and mdj groups
(13) compared to the Stata example from lecture.  My mean is about the same
(5.85).  This seems to be the same dataset

```{r DEMSV_U1}
# Model DEMSV_U1 --------------

DEMSV_U1 <- lmer(ln_ecdata ~  (1|mdj), data = DEMS, REML = TRUE, subset = facilityid == "A" & u_s == "u")
summary(DEMSV_U1)
```

## Slides 36:  DEMS V Table 3 Facility A example U2 groups

```{r DEMSV_U2}
# Model DEMSV_U2 --------------
DEMSV_U2 <- lmer(ln_ecdata ~  as.factor(group2time) + (1|mdj), data = DEMS, 
                 REML = TRUE, subset = facilityid == "A" & u_s == "u")
summary(DEMSV_U2)
```

## Slide 37:  DEMS V Table 3 Facility A example U3 groups

```{r DEMSV_U3}
# Model DEMSV_U3-------------
DEMSV_U3 <- lmer(ln_ecdata ~  as.factor(group3area) + (1|mdj), data = DEMS, 
                 REML = TRUE, subset = facilityid == "A" & u_s == "u")
summary(DEMSV_U3)
```

## Slide 41:  Regression for association with random effects

Use the snapshot data to compare estimates and SEs for a fixed effect with and
without random effects

```{r mixed model vs lm}
# mixed model vs lm -----------
# Compare estimates and SEs
# linear model:
lmfit1 <- lm(ln_nox ~ Pop_5000, data = fvc)

#summary(lmfit1)

cat("\nStandard linear model summary, coefficients then their confidence intervals\n")
coef(summary(lmfit1))

confint(lmfit1)

# mixed model
mixedfit1_REML <- lmer(ln_nox ~ Pop_5000 + (1|group_loc), data = fvc, REML = TRUE)

#summary(mixedfit1_REML)
cat("\nMixed model summary using REML, coefficients then their confidence intervals\n")
coef(summary(mixedfit1_REML))

confint(mixedfit1_REML)

```

Note (not discussed in lecture):  The fixed effect estimates will change when
there is correlation between fixed and random effect parts of the model.  Random
effects do not address residual confounding.  The most appropriate assumption is
to assume the random effects are independent of the fixed effects.  In the
following example there appears to be some correlation between the *D2Comm*
variable and the random effects, leading to a change in the coefficient estimate
for *D2Comm* between the fixed and mixed models.

```{r mixed model vs lm: 2 covars}
# mixed model vs lm: 2 covars ------------
# Compare estimates and SEs
# linear model:
lmfit2 <- lm(ln_nox ~ Pop_5000 + D2Comm, data = fvc)

cat("\nStandard linear model summary, coefficients then their confidence intervals\n")

coef(summary(lmfit2))

confint(lmfit2)

# mixed model
mixedfit2_REML <- lmer(ln_nox ~ Pop_5000 + D2Comm + (1|group_loc), data = fvc, REML = TRUE)

cat("\nMixed model summary using REML, coefficients then their confidence intervals\n")

coef(summary(mixedfit2_REML))

confint(mixedfit2_REML)

```
## Slide 42:  Regression for prediction with random effects

To predict from a mixed model, i.e. a linear model with fixed and random
effects, you MUST have the same groups as you used to fit your model.  You can't
predict on new groups.  Thus random effects models are less commonly used in
prediction problems than other kinds of variance models (e.g. geostatistical
models).

When you predict from a mixed model, you need to decide whether you want to only
predict the fixed effects, or incorporate the random effects into the
predictions.

The following shows full predictions (fixed+random) and then fixed effect only
predictions in the snapshot gradient site data for the mixed effect model with
only one covariate:

```{r ME model prediction on same data}
# mixed effects model prediction, same data -------
# Prediction incorporating fixed & random effects
p0 <- predict(mixedfit1_REML)
plot(fvc$group_loc,p0)

# Predictions of fixed effects only
# The re.form = NA option omits random effects
p1 <- predict(mixedfit1_REML, re.form = NA)
plot(fvc$group_loc,p1)

identical(p0,p1)
plot(p1,p0)
# Note the predictions with random effects are more variable, as one would expect.
summary(p1)
summary(p0)
```

When you predict in a new dataset from a mixed model, you MUST have all grouping
variables and random effects used in the model fitting.  However, if you don't
want to predict at all groups, you can set their values to NA.  For details, see
this [predict.merMod help
page](https://www.rdocumentation.org/packages/lme4/versions/1.1-19/topics/predict.merMod)
For example, to predict at one location per gradient site in the snapshot data:

```{r ME model prediction on new data}
# ME model prediction on new data--------
newdata <- with(fvc, expand.grid(Pop_5000=unique(Pop_5000), 
                                 group_loc=unique(group_loc)))
dim(newdata)
# new data, all RE
p2 <- predict(mixedfit1_REML,newdata)

# new data, level 0 (fixed effects only)
p3 <- predict(mixedfit1_REML,newdata,re.form=NA)

# explicitly specify RE model we want
p4 <- predict(mixedfit1_REML,newdata,re.form= ~(1|group_loc)) 

# verify that p2 and p4 are identical as they are supposed to be
identical(p2, p4)
plot(p3,p2)
# Note the predictions with random effects are more variable, as one would expect.
summary(p3)
summary(p2)
```


# Appendix 

```{r session.info}
#-----------------session.info: beginning of Appendix -----------------
#This allows reproducibility by documenting the version of R and every package you used.
sessionInfo()
```

```{r appendix.code, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60), include=T}

```

```{r functions, eval = TRUE}
#-----functions----

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lapply(c(lsf.str()), getAnywhere)

```
