---
title: 'Week 4 Lab:  Regression for Prediction'
author: "Lianne Sheppard for ENVH 556"
date: "Winter 2019; Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        fig_caption: yes
        toc: true
        toc_depth: 3
        number_sections: true
editor_options: 
  chunk_output_type: console
---

<!--Basic document set-up goes here  -->

```{r setup, include=FALSE}
#-------------r.setup-------------
knitr::opts_chunk$set(echo = TRUE)
```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#----------------load.libraries.pacman----
# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.  Some reasons for packages:
# knitr:  kable()
# ggplot2: part of tidyverse
# readr: part of tidyverse
# dplyr: part of tidyverse
# multcomp:  glht
# modelr:  part of tidyverse and need for add_predictions and add_residuals
# boot:  cv tools are available
# Hmisc:  describe
pacman::p_load(tidyverse, knitr, multcomp, dplyr, modelr, Hmisc)  
```

```{r read.data.from.a.web.site, eval=TRUE, echo=FALSE}
#-----read.data.from.a.web.site--------
# Download the data file from a web site if it is not already downloaded, then
# read in the file

datapath <- "Datasets"
dir.create(datapath, showWarnings=FALSE, recursive = TRUE)

snapshot.file <- "allseasonsR.rds"
snapshot.path <- file.path(datapath, snapshot.file)

# Only download the file if it's not already present
if (!file.exists(snapshot.path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 snapshot.file, sep = '/')
    download.file(url = url, destfile = snapshot.path)
}

# Output a warning message if the file cannot be found
if (file.exists(snapshot.path)) {
    snapshot <- readRDS(file = snapshot.path)
} else warning(paste("Can't find", snapshot.file, "!"))

```



# Purpose

The purpose of this lab is to use principles of “out-of-sample” assessment to
validate regression models.   We will use the snapshot data for model
validation, run a cross-validation, and write a program to more easily repeat
cross-validation procedures. You will use these tools to try to understand the
bias-variance trade-off in these data.


# Getting Started

This section gives some basic R commands for regression, prediction, and model
validation.  We will also learn how to write loops and programs.

## Set-up

* Restrict data to one season:  (fall here) 

```{r fall subset}
# fall subset ----
# Traditional base R approach
#fall <- subset(snapshot, season==2)

# Tidyverse approach
fall <- filter(snapshot, season == 2)

```

* Common model names, for later use

```{r common model names}
# common model names -------
covars_common <- c("D2A1", "A1_50", "A23_400",  "Pop_5000", "D2C", "Int_3000", 
                   "D2Comm")

```

## Commands for regression, producing AIC, plotting, and computing prediction R^2^

See also Week 3 lab for these tools and variations of applying them.

* Regression:  

```{r fall regression}
# fall regression, common model-----

summary( lm_fall <-
    lm(ln_nox ~ D2A1 + A1_50 + A23_400 + Pop_5000 + D2C + Int_3000 + D2Comm,
       data = fall))

```

* AIC and BIC:  

Note, you can provide these functions multiple model objects to facilitate comparisons of AIC or BIC across models fit on the same data.  I only show one model object here.  This has limited utility since there isn't inherent meaning in the AIC or BIC estimates.  Their value is in their comparison across fitted models.

```{r AIC and BIC}
# AIC and BIC -------
# AIC uses a default penalty of 2
AIC(lm_fall)
# BIC uses a penalty of log(n) where n is the number of observations in the
# dataset
BIC(lm_fall)

```

* Extract predictions, and plot them:  

(Note: the `fig.width` and `fig.height`
options to a chunk give local definitions for the figure height and width.)

```{r predictions with modelr, fig.width=6, fig.height=5}
# predictions with model -----------
snap2 <- snapshot %>%  
    modelr::add_predictions(lm_fall,"preds_fall")  

# Compare the observations vs. predictions for all seasons, by season 
ggplot(data = snap2, aes(preds_fall, ln_nox)) +
    geom_point() +
    coord_fixed() +
    facet_wrap( ~ seasonfac) +
    geom_abline(intercept = 0, slope = 1, color = "blue") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = "Fall model predictions vs. ln(NOx) \nby season", 
         x = "Predicted ln(NOx) (ln(ppb))",
         y = "Observed ln(NOx) (ln(ppb))",
         caption = "Best fit line is red; 1:1 line is blue")

```

* Correlations of predictions vs. observations by season:  

(Note:  these are regression-based R^2^ estimates.)

```{r season-specific correlations, eval = TRUE}
# season-specific-correlations ------
pred_summary <-
    dplyr::select(snap2,  
                  c("ln_nox", "preds_fall", "ID", "season", "seasonfac")) %>%
    group_by(seasonfac) %>%
    dplyr::summarize(r = cor(ln_nox, preds_fall),
    R2 = r ^ 2)
    
# summarize in a table
kable(pred_summary, digits = 2)
    
```

## Commands for stepwise regression (forward selection)

The R function `step()` allows stepwise regression.  It has forward, backward,
or stepwise regression search algorithms. Forward selection is a useful tool for
ordering a sequence of models based on increasing complexity.

Note:  The `step` function is available in both base R and in the `MASS`
package.  While we won't use it in this lab, the function `addterm()` from the
`MASS` package allows the addition of the next best term to a model.  This
allows use of the F-test or a Chi-square test as alternatives to AIC which is implemented below.

* **Model Set-up**
First define the smallest starting model (the *null* model) and the largest
possible model we will consider (the *full* model). These define the scope.  The
function `step()` has options to specify the direction of the model selection
(we will only use forward selection or `direction = "forward"`), the amount of
output (we'll use `trace = 0` to omit printing any output) and the number of
models fit (the default number of models fit, or `steps`, is 1000).  `step()`
minimizes the AIC and the option `k` controls the degree of penalty.  Since for
our purposes here we don't want to stop early, we use `k = 0` which seems to not
impose a stopping criterion in our dataset.  More investigation is needed to
understand exactly how the `k` option operates w.r.t. stopping the alorithm.


```{r forward selection set-up for fall snapshot}
# forward selection set-up for fall snapshot ------
# define the smallest model of interest, an intercept only model here
null <- lm(ln_nox ~ 1, data=fall)
#null
    
# create the largest possible model, a full model that includes all of the
# predictor variables in the dataset.
# Steps:
# 1: get the list of all the potential covariates of interest from the dataset:
covars_all <- names(fall[12:74])

# 2: now turn this into a formula for the full model in stepwise regression:
full <- as.formula(paste("ln_nox ~ ", paste(covars_all, collapse= "+")))
#full

```   

* **Forward Selection**

The first model listed in the command (the "object"), is the initial model used to start the search.  In forward regression we want to start with the smallest model of interest to us, here the intercept only model.  The scope gives the entire range of models by also incorporating the largest possible model.  

```{r fitting forward selection}
# forward selection model------
# Note:  k=0 appears to put no restriction on the forward selection and doesn't
# stop until the full model is incorporated. 
# Using k=2 is comparable to standard AIC.
# Using log(n), where n is the number of observations, is comparable to BIC.
forwardreg_fall <- step(null, scope=list(lower=null, upper=full), trace = 0,
                        direction="forward", k=0)

# save the ordered list of names for later use, dropping the intercept
covars_forward <-       names(forwardreg_fall$coefficients)[-1]

```

```{r explore stepwise results, eval=FALSE}
# explore results from forward stepwise-----
# structure of the object
str(forwardreg_fall)

# class of the object
class(forwardreg_fall)

# gives the names of the coefficients for use in other work.  Should assign a
# name to it for later use.
names(forwardreg_fall$coefficients)

# print a few other things, preceded by what they are in the output (by using
# the cat() command to print text in the output.)
cat("AIC:  ", "\n")
forwardreg_fall$anova$AIC
cat("Deviance:  ", "\n")
forwardreg_fall$anova$Deviance
forwardreg_fall$anova$Step
#forwardreg_fall$anova$Df

```


## Manual cross-validation and summarization

Note:  We are summarizing the cross-validation results overall.  This overall summary is one
valid way to do the summary.  Other approaches summarize the results in each
cluster (or cross-validation group) separately and average these.  (See e.g. the *Introduction to Statistical Learning* textbook: 
[ISL](http://www-bcf.usc.edu/~gareth/ISL/) for guidance using this approach.)


```{r a manual CV}
# manual CV NEW (with better practices) ------

# create a variable to store CV predictions 
# Note: it is better practice to not grow objects as we have in the commented
# out statement thta follows.
#cv_pred <- NULL
cv_pred <- data.frame(fall, preds = numeric(dim(fall)[1]))

# loop over the 10 clusters
for (i in 1:10){
    # define the current cluster variable as a logical vector
    is_cluster <- cv_pred$cluster == i
    
    # fit the linear model to the training set by omitting cluster i 
    # Note the better practice to negate logical vectors with "!" and not "-".
    CV_lm <-
        lm(ln_nox ~ D2A1 + A1_50 + A23_400 + Pop_5000 + D2C + Int_3000 + D2Comm,
           data = cv_pred, subset =!is_cluster)
  
    # save the predictions and observations in cv_pred for the current cluster
    # (test set)
    cv_results <- fall %>%
        modelr::add_predictions(CV_lm, "preds") 
    #%>%
        #filter(cluster == i) %>%
        #dplyr::select(ID, cluster, ln_nox, preds)
    
    # bind the results together to get back the full dataset    
    cv_pred$preds[is_cluster] <- cv_results$preds[is_cluster]
}

# now get the MSE, RMSE, MSE-based R2
# 
# mean of observations
ln_nox_avg <- mean(cv_pred$ln_nox)

# MSE of predictions
MSE_pred <- mean((cv_pred$ln_nox - cv_pred$preds)^2)

# MSE of observations (for R2 denominator)
MSE_obs <- mean((cv_pred$ln_nox - ln_nox_avg)^2)

# print the results, rounded
#paste("RMSE:  ", round( sqrt(MSE_pred), 2))
#paste("MSE-based R2:  ", round( max(1 - MSE_pred/MSE_obs, 0), 2))
# print the results not rounded
paste("RMSE:  ", sqrt(MSE_pred))
paste("MSE-based R2:  ", max(1 - MSE_pred/MSE_obs, 0))  

```


## Functions for cross-validation and MSE 

### Tips on writing functions

From the *R for Data Science* [R4DS](https://r4ds.had.co.nz/index.html) chapter on functions, [Chapter 19](https://r4ds.had.co.nz/functions.html), the
reason to write functions is to automate repetitive tasks and avoid copying and pasting.  This eliminates the need to update the code in multiple places when you make a change and reduces your chances of making mistakes.  A simple rule of thumb is to write a function whenever you have or will need at least three copies of a block of code.

Here are the 3 basic steps to writing a function as described in *R4DS*:

1. Pick a name for the function.
2. List the inputs or *arguments* to the function inside the function call.
3. Put code to accomplish what you want in the body of the function.

Here is a list of twelve best practices for writing functions graciously provided by Brian High:  

1. Use a function when you need to repeat a block of code, instead of simply 
   copying and pasting the same block of code. I.e. "do not repeat yourself".
2. If the code would need to be slightly different with each execution, 
   allow for those differences in the function's argument (parameter) list.
3. All inputs needed for the function should be in the argument list. Do not
   refer to variables within the function which are defined outside of it, but
   were not included in the parameter list. I.e., "make no assumptions".
4. Define default values in the argument list if it makes sense to do so. 
5. The order of arguments should be from most to least essential. If the 
   function operates on a data frame, for example, it should be the first 
   argument. Put arguments with defined defaults at the end, so the function 
   user can safely omit providing values for those arguments if they prefer to 
   use the defaults you have defined for them.
6. Use argument names as consistently as possible. Refer to functions you 
   commonly use for examples of consistently named arguments.
7. The function should return a single value or object (e.g., a vector, 
   dataframe, list, etc.).
8. Use the return() function at the end of your function to return this object 
   explicitly. Otherwise, the value of the last expression of the function will 
   be returned implicitly. You will see both methods in practice, but the use 
   of return() is preferred as it makes it very clear what is being returned.
9. Comment your function to make it clear what it expects for input and what
   it returns as output
10. Write the function to be as resuable (i.e., flexible, generic) as practical.
11. When practical, validate the inputs (argument list) for expected data type 
   and appropriate range if values to avoid strange results if the function is 
   misused.
12. Avoid huge monolithic functions that perform all sorts of tasks. Separate 
    steps into separate functions when it makes sense to do so. This will make 
    your code more reusable, flexible, easier to understand, and therefore 
    easier to debug.


### Our first function:  getMSE

To start our learning process, first we write a function to estimate the MSE and
MSE-based R^2^.  We call it `getMSE`.  Note that in addition to its name and
arguments, there are key formatting details that are required in defining
functions in R.  In particular, pay attention to the arguments inside the
parentheses in the function command, and the use of curly brackets.

```{r define getMSE}
# define getMSE function -----
# This is a function to get the MSE, RMSE, MSE-based R2
getMSE <- function(obs,pred) {
    # obs is the outcome variable
    # pred is the prediction from a model
    # 
    # mean of obs
    obs_avg <- mean(obs)
    
    # MSE of obs (for R2 denominator)
    MSE_obs <- mean((obs-obs_avg)^2)
    
    # MSE of predictions
    MSE_pred <- mean((obs - pred)^2)
    result <- c(sqrt(MSE_pred),
                max(1 - MSE_pred / MSE_obs, 0))
    names(result) <-  c("RMSE", "R2(MSE-based)")
    # it is better practice to use an explicit return statement
    # the alternative is to just output the object:
    #result
    return(result)
}

```

Now test our function, using the manual CV results we obtained:

```{r test getMSE}
# test getMSE -----
getMSE(cv_pred$ln_nox,cv_pred$preds)

```

### A function to do cross-validation:  do_CV

Now convert our cross-validation to a function:

```{r CV function}
# CV function -------
do_CV <- function(data, group = "group", formula) {
    # In writing functions, it is good practice to:
    # 1. put the dataset first in the function definition
    # 2. put most important arguments first, followed by the specific ones
    # 3. can include defaults in the function definitions
    # 
    # Arguments:
        # data is the data frame
        # group is the grouping variable (in the data frame)
        # formula is the formula to pass to lm
        # the function returns the dataset with a new variable called cvpreds
        # appended to the end; these are the , out-of-sample predictions

    # create a variable to store CV predictions 
    #CV_pred <- NULL
    # This better practice version appends the new variable to the input data
    # frame
    CV_pred <- data.frame(data, cvpreds = numeric(dim(data)[1]))
                          
    # get the number of distict groups or clusters
    # Note the use of "[[ ]]" rather than "$" because group is input in the
    # function call as a quoted variable
    k <- length(unique(data[[group]]))
    
    # loop over all the groups or clusters
    for (i in 1:k){
        # define the current group variable as a logical vector
        # again using "[[ ]]" instead of "$"
        is_group <- data[[group]] == i
        
        # subset the data to create the training set by omitting cluster i
        # Note: better to use "!" rather than "-" for negating a logical vector
        training <- data[!is_group, ]
        
        # fit the linear model to the training set
        CV_lm <- lm(formula = formula, data = training)
  
        # save the predictions in a temporary dataset with all observations
        # (training + test sets), tempCV, then restrict updating into CV_pred to
        # the current group (test set)
        tempCV <- data %>% 
            modelr::add_predictions(CV_lm, "cvpreds") 
        # only update CV_pred dataset, cvpreds variable, for the test data rows:
        CV_pred$cvpreds[is_group] <- tempCV$cvpreds[is_group]
    }
    # return the dataset
    return(CV_pred)
}

```

Now test the `do_CV` function:

```{r test do_CV}
# test do_CV------

# first create the formula from a vector of names
common_model_formula <-
    as.formula(paste("ln_nox ~ ", paste(covars_common, collapse = "+")))

# Note: the do_CV function assumes the group variable is part of the input dataset. 

temp <- do_CV(data = fall, group = "cluster", formula = 
                  common_model_formula)


# Check to see if result is identical with manual method (done earlier)
identical(cv_pred$preds, temp$cvpreds)


```

And review the CV results:

```{r review CV results}
# review CV results -------
# now check the results from the CV function
# Summarize with getMSE
getMSE(temp$ln_nox, temp$cvpreds)

# now look at the scatterplot, observations on the x axis as we typically do for
# considering out-of-sample predictions
ggplot(data = temp, aes(ln_nox, cvpreds)) +
    geom_point() +
    coord_fixed() +
    geom_abline(intercept = 0, slope = 1, color = "blue") +
    labs(title = "Fall model predictions vs. ln(NOx) \nCross-validated", 
         x = "Observed ln(NOx) (ln(ppb))",
         y = "Predicted ln(NOx) (ln(ppb))",
         caption = "1:1 line is blue")

```

## Create 10 CV groups in random order

```{r generate groups}
# generate groups --------
# set the seed to make reproducible
set.seed(283)

# first generate a vector with values 1:10 with as many observations as the dataset
CVgrp <- rep(1:10, length.out = dim(fall)[1])

# now give it a random order
# replace = FALSE in sample just reorders the variable
CVgrp <- sample(CVgrp, replace = FALSE)

# now append it to the fall data frame
# Base R version
#fall <- cbind(fall, CVgrp)

# Tidyverse version
fall <- mutate(fall, CVgrp = CVgrp)

```

Now test the CV on random groups.  Note that these out-of-sample results should look better.  Why is that the case? (This is a homework challenge question.)

```{r CV results on random group}
# CV results random group -------
# fit the CV with the random groups
temp2 <- do_CV(data = fall, group = 'CVgrp', formula = common_model_formula)

# now check the results from the CV function
# Summarize with getMSE
getMSE(temp2$ln_nox, temp2$cvpreds)

# now look at the scatterplot, observations on the x axis as we typically do for
# considering out-of-sample predictions
ggplot(data = temp2, aes(ln_nox, cvpreds)) +
    geom_point() +
    coord_fixed() +
    geom_abline(intercept = 0, slope = 1, color = "blue") +
    labs(title = "Fall model predictions vs. ln(NOx) \nCross-validated with random groups", 
         x = "Observed ln(NOx) (ln(ppb))",
         y = "Predicted ln(NOx) (ln(ppb))",
         caption = "1:1 line is blue")

```

## Bias-variance trade-off analysis 

Assignment Request:	Use the order of entry into the stepwise to sequentially
complete the following computations.  For models with one up to the maximum
number of selected variables:  

1. Use the full dataset to obtain in-sample estimates of the RMSE and R2.  
2. Estimate predicted values using cross-validation.  
3.  Compute out-of-sample RMSE and MSE-based R2 estimates.  
 
Steps to coding this:  

* compute/retrieve the vector of names output from the forward selection model

* loop over the maximum number of covariates from forward selection

* create a formula that is the linear combination of the first i terms of the forward selection model

* summarize the in-sample prediction MSE and R2

* cross-validate this and summarize the out-of-sample MSE and R2

* store the key set of results, plus the number of covariates in the model and possibly the variable added to use in plotting

* use ggplot to make the bias-variance trade-off plot


```{r Model order and CV, echo = TRUE, warning = FALSE}
# Model order and CV ----

#get the length of the vector of names from forward selection
n_covars <- length(covars_forward)

# set up an object to store the results in 
# this is better practice
res <-
    tibble(
        n_pred           = numeric(n_covars),
        covar            = character(n_covars),
        in_RMSE           = numeric(n_covars),
        in_R2            = numeric(n_covars),
        out_RMSE          = numeric(n_covars),
        out_R2           = numeric(n_covars)
    )
# here is the base R approach.  Note need to use stringsAsFactors = FALSE
#res <-
#    data.frame(
#        n_pred           = numeric(n_covars),
#        covar            = character(n_covars),
#        in_RMSE           = numeric(n_covars),
#        in_R2            = numeric(n_covars),
#        out_RMSE          = numeric(n_covars),
#        out_R2           = numeric(n_covars),
#        stringsAsFactors = FALSE
#    ) 

# loop over n_covars
for (i in 1:n_covars) {
    # assign first few variables
    res$n_pred[i] <- i
    res$covar[i] <- covars_forward[i]
    
    # define the formula, updated to add a term each time
    fmla <- as.formula(
        paste("ln_nox ~ 1 + ", paste(covars_forward[1:i], collapse = "+"))) 
    
    # in-sample model and estimates
    in_model <- lm(fmla, data = fall) 
    res$in_RMSE[i] <- sqrt(mean(in_model$residuals^2))
    res$in_R2[i] <- summary(in_model)$r.squared
    
    # out-of sample model and estimates
    out_ests <- do_CV(fall, group = 'cluster', fmla)
    out_results <- getMSE(out_ests$ln_nox, out_ests$cvpreds)
    res$out_RMSE[i] <- out_results[1]
    res$out_R2[i] <- out_results[2]
    
    }

# the data frame res should have all pieces for ggplot


```

### Now plot the bias-variance trade-off

Note: the precipitous drop in R2 in the out of sample assessment is best explained by major overfitting of this model after ~ 30 covariates.

```{r bias-var separate plots}
# bias-var separate plots ------

# first reshape the data to take advantage of ggplot features
res.reshaped <- res %>% 
    gather(key = variable, value = value, -n_pred, -covar) %>% 
    separate(col = variable, into = c('Source', 'Estimate'), sep="_") %>% 
    spread(key = Estimate, value = value)

# R2 plot
# plots with both model based (in sample) R2 and CV generated R2
plot.R2 <- ggplot(res.reshaped) +
    geom_point(aes(x = n_pred, y = R2, color = Source)) + 
    geom_line(aes(x = n_pred, y = R2, color = Source)) +
    xlab("Model Complexity (# of terms)") +
    ylab(bquote(bold(R ^ 2))) +
    scale_x_continuous(breaks = c(seq(0, 63, 5))) +
    theme_bw() +
    theme(axis.text = element_text(face = "bold"),
          axis.title = element_text(face = "bold")) +
          scale_color_discrete(name = "Prediction\nSource",  
                labels = c("In-sample", "Out-of-sample"))

print(plot.R2)


# plot with both model based (in sample) RMSE and CV generated RMSE

plot.RMSE <- ggplot(res.reshaped) +
    geom_point(aes(x = n_pred, y = RMSE, color = Source)) + 
    geom_line(aes(x = n_pred, y = RMSE, color = Source)) +
    xlab("Model Complexity (# of terms)") +
    ylab(bquote(bold(RMSE))) +
    scale_x_continuous(breaks = c(seq(0, 63, 5))) +
    theme_bw() +
    theme(axis.text = element_text(face = "bold"),
          axis.title = element_text(face = "bold")) +
          scale_color_discrete(name = "Prediction\nSource",  
                labels = c("In-sample", "Out-of-sample"))

print(plot.RMSE)

```

```{r bias-var combined plots}
# bias-var combined plots ------
# TODO:  Some clean-up details should be addressed to help the presentation.

# plots with both model-based (in sample) and CV generated MSE and R2

# first reshape the data to take advantage of ggplot features
# No longer want the spread used above
res.reshaped <- res %>% 
    gather(key = variable, value = value, -n_pred, -covar) %>% 
    separate(col = variable, into = c('Source', 'Estimate'), sep="_") 


plot.both <- ggplot(res.reshaped) +
    geom_point(aes(x = n_pred, y = value, color = Source)) + 
    geom_line(aes(x = n_pred, y = value, color = Source)) +
    xlab("Model Complexity (# of terms)") +
    ylab("") +
    scale_x_continuous(breaks = c(seq(0, 63, 5))) +
    facet_grid(Estimate ~ ., scales = "free_y") +
    theme_bw() 

print(plot.both)

```

# Practice Session

This section covers basic practice to be completed during the lab.   We are
going to use the snapshot data described in Mercer et al 2011, discussed in
class, and used last week.  It can be found on the class website.  Note that in
this lab we are treating the errors are independent and identically distributed
even though this assumption is not correct in these data.  (This is the same
assumption typically made in LUR modeling.)

Perform the following tasks:

1. Determine the R project you will use.
2. Explore the dataset a bit, focusing on one particular season.  Make sure you
have some basic understanding of the outcome variable (*ln_nox*), the CV grouping
variable (*cluster*), and the large number of covariates you can use in your prediction modeling.   In
this lab you should restrict your analysis to one season (e.g. fall for the practice session; winter for homework).
3. Fit the model for one of the seasons given in Table 4 of Mercer et al. Make
note of these in-sample estimates of R^2^ and RMSE.
4. Try to manually cross-validate this model using the code given above.
Compare the CV R^2^ and RMSE to your in-sample estimates.  (Note:  Use the cluster
variable in the dataset to define your CV groups.)
5. Use the cross-validation function and repeat your cross-validation analysis.
(If you use the same groups, you should get the same results as in the previous
step.)
6. Make a scatterplot comparing *ln_nox* (the observed dependent variable) on the
x-axis with the cross-validated predictions on the y-axis.  Add the 1:1 line to
your plot. (If you also want to show the best-fit line, you’ll need to put the
predictions on the x-axis rather than the y-axis.)
7. Create your own version of the bias-variance trade-off plot shown in class
using the following steps:
    a. Do a forward stepwise regression of *ln_nox* on a set of plausible variables
with a lax entry criterion (`k=0` in the `step()` function.  (You may restrict
your attention to the list in the forward selection model example given above.)
See the code to learn how to keep track of the order the variables were added.
    b. Use the order of entry into the stepwise to sequentially complete the
following computations.  For models with one up to the maximum number of
selected variables:
        i. Use the full dataset to obtain in-sample estimates of the RMSE and R^2^.
        ii. Estimate predicted values using cross-validation.
        iii. Compute out-of-sample RMSE and MSE-based R^2^ estimates.
    c. In a table or figure(s), summarize the number of variables in each model
along with the R^2^ and/or RMSE estimates from CV and the training data.  If you
choose to show your results in a plot, put the number of variables in the model
on the x-axis vs. the R^2^ or RMSE estimates on the y-axis.  Distinguish the two
kinds of estimates on your plot.  If you choose to show your results in a table,
also include a column for the variable name of the variable added.

# Homework Exercises

1. Write a brief summary of the purpose of the lab and your approach.  Then present your results:

    a.  Describe the results (with appropriate displays in table(s) and/or
    figures(s)), and
    
    b.  Discuss the insights you have obtained from your analyses, both of the
    training data alone and after cross-validation. In your discussion, comment
    on how your in-sample and cross-validated MSE-based R^2^ estimates compare.
    
2. **Extra credit**. Present one or both of the following results in your write-up:

    a.  Repeat the exercise using randomly defined CV groups that ignore the
    gradient clusters.
    
    b.  Repeat the exercise using yet another different set of CV groups, either
    chosen randomly, or spatially, or based on some other criterion.



# Appendix 

```{r session.info}
#-----------------session.info: beginning of Appendix -----------------
#This allows reproducibility by documenting the version of R and every package you used.
sessionInfo()
```

```{r appendix.code, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60), include=T}

```

