---
title: "Week 3 Lab:  Regression for Association"
subtitle: "Companion document with more advanced coding methods"
author: "Lianne Sheppard & Chris Zuidema for ENVH 556"
date: "Winter 2021; Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        df_print: "paged"
---

```{r setup, include=FALSE}
#-----setup-----

# set knitr options
knitr::opts_chunk$set(echo = TRUE)

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.  Some reasons for packages:
pacman::p_load(knitr, dplyr, ggplot2, egg, tidyr, broom, purrr, stringr)

```

```{r directory.organization.read.data, echo=FALSE, warning=FALSE}
#-----directory organization and read data-----

# specify working directory
project_path <- getwd()

# create "Datasets" directory if one does not already exist    
dir.create(file.path(project_path,"Datasets"), showWarnings=FALSE, recursive = TRUE)

# specify data path
data_path <- file.path(project_path,"Datasets")

# specify the file name and path
file_name <- "allseasonsR.rds"
file_path <- file.path(data_path, file_name)

# Download the file if it is not already present
if (!file.exists(file_path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 file_name, sep = '/')
    download.file(url = url, destfile = file_path)
}

# Output a warning message if the file cannot be found
if (file.exists(file_path)) {
    snapshot <- readRDS(file_path)
} else warning(paste("Can't find", file_name, "!"))

# remove temporary variables
rm(url, file_name, file_path, data_path)

```


# Purpose

This document is a companion to the `Week3Lab.Rmd`. The purpose is to provide
additional and somewhat more sophisticated coding methods to supplement the
approaches in Lab 3 that you may find useful for your own work. Our coding goals
for `Week3Lab.Rmd` were clarity and transparency to support learning about
regression. Here we explore techniques that may scale well (for example to many
regression models), reduces code repetition, and reduces the amount of "hard
coding" in the analysis, but is less explicit and may obscure what is going on
"behind the scenes" a bit. As with `Week3Lab.Rmd`, this document will use the
ACT snapshot data.


# Preliminaries

These preliminary steps are the same as `Week3Lab.Rmd` and are included for
completeness.

## Prepare snapshot dataframe

```{r prep.snapshot}
#-----prep snapshot-----

# transform snapshot dataframe to tibble and modify season factor
snapshot <- snapshot %>% 
  as_tibble() %>% 
  mutate(seasonfac = factor(str_remove(seasonfac, "[:digit:]"),   
         levels = c("Summer", "Fall", "Winter")) )

# create season variable (to be used throughout analysis)
seasons <- unique(snapshot$seasonfac)

```

## Summarize *ln_nox* by season

```{r summarize.ln_nox}
#-----summarize ln_nox-----

# frequencies only
summary(snapshot$seasonfac)

# descriptive statistics
snapshot_summary <- snapshot %>%
    group_by(seasonfac) %>%
    summarise(N = n(),
              Mean = mean(ln_nox), 
              SD = sd(ln_nox),
              GM = exp(mean(ln_nox)), 
              GSD = exp(sd(ln_nox)), 
              .groups = "drop") %>% 
    mutate_if(is.double, round, digits = 2)

# show summary
snapshot_summary

```

## Season-specific boxplots

```{r season-specific boxplots}
#-----season-specific boxplots-----

ggplot(snapshot, aes(x = seasonfac, y = ln_nox, fill = seasonfac)) +
    geom_boxplot() +
    labs(x = "Season", 
         y = "ln(NOx) (ln(ppb)", 
         color = "Season", 
         fill = "Season") + 
    theme_article()

```

# Regression analysis

`Week3Lab.Rmd` takes a bit more of a manual approach to regression modeling,
specifying each of the seasonal models separately and assigning each model fit
to its own object. This is clear and easy to follow, but you can see how this
would quickly become a burdensome exercise in copy/paste and repeated code -
imagine if we had 20 seasons instead of 3. In this section, we explore the
general nesting workflow, where data is grouped and modeling steps occur
"within" a dataframe. Aside from managing analysis steps with groups
programmatically, these functions package regression output in dataframes that
are `ggplot` friendly - minimizing the data wrangling required before plotting.

You can find more information about the nested workflow here:

  * Tidyverse ["Nested Data"](https://tidyr.tidyverse.org/articles/nest.html) 
  * Bookdown ["Many Models"](https://r4ds.had.co.nz/many-models.html)

```{r regression for association}
#-----regression for association-----

# Common model (Table 4), season-specific LUR from Mercer et al, 2011

# specify the "common model"
frml <- as.formula(ln_nox ~ D2A1 + A1_50 + A23_400 + Pop_5000 + D2C + Int_3000 + D2Comm)

# create tibble with nested seasonal models
season_mods <- snapshot %>%
  
  # group by season, then "nest"
  group_by(seasonfac) %>% 
  nest() %>% 
  
  # fit regression model for each season and generate summary, glance, augment,
  # and in-sample prediction outputs.
  mutate(model = map(data, ~lm(frml, data = .x)),
         summary = map(model, summary), 
         glanced = map(model, glance),
         augmented = map(model, augment),
         preds = map2(model, data, 
                      ~as.data.frame(predict(.x, .y, interval = "prediction")) )
         
         )

# show tibble
season_mods
```

You can see `season_mods` is a grouped tibble with list-columns for each of the
non-grouping variables ("seasonfac"). Next we'll demonstrate how to access these
outputs separately to show what outputs the `glance()` (model summaries) and
`augment()` (information about each observation like predictions and residuals)
functions produce. But, you can also incorporate the `unnest()` step into the
pipeline in the previous chunk. We showed the `summary()`, `glance()`,
`augment()`, and `predict()` functions above, but you can include or remove
functions to suit your needs.

```{r unnest}
#-----unnest-----

# unnest glanced object - `glance()` provides model summaries
unnest(season_mods, glanced)

# unnest augmented dataframe for ".fitted" and ".resid" values. Unfortunately, 
# the prediction intervals are not provided with `augment()`. We'll now assign 
# this output to the `season_mods` object, but this could have been accomplished 
# by continuing the pipeline in the previous chunk. We split up this pipeline 
# just to demonstrate how to access the different objects produced in the nest 
# workflow.
season_mods <- unnest(season_mods, augmented)

# show augmented season_mods object
season_mods
```

# Regression predictions

## Output from `nest()` workflow

In the previous chunks we calculated in-sample predictions with `augment()`. The
nesting workflow makes plotting those predictions easy:

```{r plot.obs.pred.in.sample, message=FALSE}
#-----plot obs pred in sample-----

# calculate range of data
r <- range(season_mods[, c("ln_nox", ".fitted")])

# plot observations vs predictions for each season
ggplot(season_mods, aes(x = .fitted, y = ln_nox)) + 
  geom_point(shape = "o") + 
  facet_wrap(~seasonfac, nrow = 1) + 
  lims(x = r, y = r) +
  coord_fixed() +
  geom_abline(intercept = 0, slope = 1, color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "In-sample Predictions", 
       x = "In-sample predicted ln(NOx) (ln(ppb))",
       y = "Observed ln(NOx) (ln(ppb))"
       ) + 
  theme_bw()
  
```

## Calculating predictions programmatically

In `Week3Lab.Rmd` we specified each of the out-of-sample predictions we were
interested in, for example:

```

fall_preds_from_summer <- 
  predict(lm_summer, snapshot[snapshot$season == 2,], interval = "prediction")

```

This is quite clear what's happening, but again, it requires us to specify the
model, the data subset, and the object name for each out of sample predictions
of interest.

Next we'll demonstrate how to calculate the season-specific predictions in the
dataset for all seasons programmatically. While `augment()` conveniently gave us
"in-sample" predictions, the following approach will give us both "in-sample"
predictions (e.g. using the summer model to make predictions in winter) and
"out-of-sample" predictions (e.g. using the summer model to make predictions in
winter). The extra step involved with this approach is selecting the output of
interest, for example, the "Fall_" columns from the Summer "seasonfac" rows.

```{r predictions.with.dplyr, message=FALSE}
#-----predictions with dplyr-----

# first, extract models from earlier regression procedures for convenience, 
# setting the names of the models to match with their season
models <- season_mods[["model"]] %>% 
  unique() %>% 
  setNames(unique(season_mods[["seasonfac"]]))

# use lapply and an anonymous function to loop through the models
allseas_both <- lapply(names(models), function(i){
  
  # add the predictions, intervals, and residuals for each season
  predict(models[[i]], snapshot, interval = "prediction") %>% 
    as_tibble() %>% 
    mutate(resid = snapshot[["ln_nox"]]-fit) %>%
    
    # rename columns to correspond to the season model
    rename_all(function(x){paste0(i, "_", x)})
  
  }) %>% 
  
  # bind list of dataframes together
  reduce(cbind) %>% 
  
  # add to snapshot dataframe
  bind_cols(snapshot, .)

# inspect predictions and residuals
allseas_both %>% 
  select(seasonfac, ln_nox, matches("_fit|_lwr|_upr|_resid")) %>% 
  mutate_if(is.double, round, 3)

```

Assessing the the quality of the predictions, as we did in `Week3Lab.Rmd`
requires us to specify the data of interest since we've produced all 
predictions together:

```{r prediction.assessment}
#-----prediction assessment

# summer in-sample
paste("summer in-sample R2:  ",
with(allseas_both %>% filter(seasonfac == "Summer"), 
     round(cor(ln_nox, Summer_fit)^2, 3)) )

# summer out-of-sample (from winter model)
paste("summer out-of-sample R2 (from winter model):  ", 
with(allseas_both %>% filter(seasonfac == "Summer"), 
     round(cor(ln_nox, Winter_fit)^2, 3)) )

```
## Plot predictions

We can define a flexible plot function to prevent repeating our plotting code:

```{r define.plot.function}
#-----define plot function-----

# make plot function
plot_func <- function(df = allseas_both, x_var, y_var, seas, 
                      x_lab, y_lab, plot_title){

  # create function variable names with `enquo`
  x <- enquo(x_var)
  y <- enquo(y_var)
  
  # get range of plots
  r <- df %>% 
    filter(seasonfac == seas) %>% 
    select(Summer_fit, Fall_fit, Winter_fit, ln_nox) %>% 
    range()
  
  # make plots
  df %>%
    filter(seasonfac == seas) %>%
    ggplot(aes(x = !!x, y = !!y)) +
    geom_point(shape = "o", alpha = 0.8) +
    lims(x = r, y = r) +
    coord_fixed() +
    geom_abline(intercept = 0, slope = 1, color = "blue") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = plot_title, 
         x = x_lab,
         y = y_lab
         ) + 
    theme_bw()

}

```


With the `allseas_both` dataframe (specified by default) we can plot in- and
out-of-sample predictions like we did in `Week3Lab.Rmd` with our function:

```{r summer.in.and.out.sample, message=FALSE}
#-----summer in and out of sample-----

# plot the predictions vs. data for summer, both in-sample and out of sample

# in-sample plot
p_in <- plot_func(x_var = Summer_fit, y_var = ln_nox, seas = "Summer", 
                  x_lab = "In-sample predicted ln(NOx) (ln(ppb))", 
                  y_lab = "Observed ln(NOx) (ln(ppb))", 
                  plot_title = "In-sample")
 
# out-of-sample plot (predictions from the winter model)
p_out <- plot_func(x_var = Winter_fit, y_var = ln_nox, seas = "Summer", 
                   x_lab = "Out-of-sample predicted ln(NOx) (ln(ppb))", 
                   y_lab = "Observed ln(NOx) (ln(ppb))", 
                   plot_title = "Out-of-sample (from winter model)" )

# combine plots
ggarrange(p_in, p_out, ncol = 2, 
          top = "Summer Model ln(NOx) Predictions", 
          bottom = "Best fit line is red; 1:1 line is blue")

```

Here is an alternate approach for plotting all predictions for a season
simultaneously using `facet wrap()`:

```{r plot.all.preds, message=FALSE}
#-----plot all predictions-----

# create temporary dataframe for ggplot
temp <- allseas_both %>% 
  select(seasonfac, ln_nox, contains("_fit")) %>% 
  filter(seasonfac == "Summer") %>%
  pivot_longer(cols = contains("_fit"), names_to = "season_fit", 
               values_to = "fit") %>% 
  mutate(season_fit = factor(str_remove(season_fit, "_fit"), 
                             levels = c("Summer", "Fall", "Winter")) )

# get range for plots
  r <- temp %>% 
    select(fit, ln_nox) %>% 
    range()
  
# make plot
pred_plot <- ggplot(data = temp, aes(x = fit, y = ln_nox) ) +
    geom_point(shape = "o", alpha = 0.8) + 
    facet_wrap(~season_fit, nrow = 1) +
    lims(x = r, y = r) +
    coord_fixed() +
    geom_abline(intercept = 0, slope = 1, color = "blue") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = "Predictions with Summer Model", 
         x = "Predicted ln(NOx) (ln(ppb))",
         y = "Observed ln(NOx) (ln(ppb))"
         ) + 
    theme_bw()

# show plot
pred_plot
```


# Appendix

```{r session.info}
#-----session information----

# print R session information
sessionInfo()

```

```{r code.appendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE, , include=TRUE}
#-----code appendix----
```

```{r functions, eval = TRUE}
#-----functions----

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lapply(c(lsf.str()), getAnywhere)

```

