---
title: "Week 9 Lab:  Time Series Data and Analysis Strategies"
author: "Chris Zuidema and Elena Austin for ENVH 556"
date: "Winter 2021; Updated `r format(Sys.time(), '%B %d, %Y')`"
output: 
    html_document:
        df_print: "paged"
        fig_caption: yes
        toc: true
        toc_depth: 3
        number_sections: true
---

```{r setup, include=FALSE}
#-----setup-----

# set knitr options
knitr::opts_chunk$set(echo = TRUE)

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
    rm(res)
}

```

```{r load.libraries, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.
pacman::p_load(knitr, dplyr, tidyr, readr, stringr, ggplot2, hms, 
               lubridate, purrr, tsibble, feasts, slider, ggmap, ggrepel)

```

```{r directory.organization.and.read.data, echo=FALSE, warning=FALSE}
#-----directory organization and read data-----

# specify working directory
project_path <- getwd()

# specify data directory
data_dir <- file.path(project_path, "Datasets", "mobile_monitoring")

# Download files if not already present
if (!dir.exists(data_dir)) {
    
  # create "Datasets" and "mobile_monitoring" directories if they do not already
  # exist
  dir.create(file.path(project_path, "Datasets", "mobile_monitoring"), 
             showWarnings = FALSE, recursive = TRUE)

  # specify URL of zip file
  url <- "https://faculty.washington.edu/sheppard/envh556/Datasets/mobile_monitoring/Jul19MOV-UP.zip"
  
  # name destination file
  dest <- file.path(data_dir, "Jul10MOV-UP.zip")
  
  # download zip
  download.file(url = url, destfile = dest)
  
  # unzip file
  unzip(zipfile = dest, exdir = data_dir)

  # delete zip files (optional)
  unlink(dest)
  unlink(str_subset("__MACOSX", list.files(data_dir)), recursive = TRUE)
  
  }
```

<!-- TODO:  THINGS TO ADD/MODIFY/REFRAME IN LAB:
LS:  I think this can be deleted unless Elena sees something more to add

# Time alignment before combining dataset
(all instruments are 1-sec interval currently...) -- LS:  Sufficient now?  I think so

# Spatial Smoothing
average by distance covered? -- LS:  Skip

# Removing the Influence of Local Exhaust Plumes
rolling 25th percentile method used by Choi et al. (2012)
outlier resistant statistics such as the median -- LS:  I think we have enough examples of rolling background

# Characterizing Background
Another task in mobile monitoring studies is to characterize "background"
concentrations. LS:  I think this is done

# Lab replicates concepts in:
[Brantley et al., 2014](https://amt.copernicus.org/articles/7/2169/2014/)
[full text](https://amt.copernicus.org/articles/7/2169/2014/amt-7-2169-2014.pdf)

-->


# Introduction and Purpose

The purpose of this lab is to practice working with time series data, develop
facility in importing and merging data of varying time scales, smoothing
approaches, and some analysis tools to reduce the complexity of these data. In
addition, as with all the labs in this course, you should focus on writing up a
coherent lab report that addresses the scientific context and scientific
questions.

In this lab exercise we will use data collected as part of the MOV-UP mobile
monitoring campaign. This data was collected on 1-second and 10-second time
scales. The lab will walk you through the steps of importing the raw instrument
data. The MOV-UP study has collected data from the following instruments: 

  * GPS tracker
  * particle counter (P-Trak, model 8525, TSI Inc., MN)
  * condensation particle counter (CPC, model 3007, TSI Inc., MN)
  * black carbon aethelometer (microAeth AE51) 
  * particle sizing spectrometer (NanoScan)
  * carbon dioxide (CO<sub>2</sub>) (SenseAir K-30-FS)

The full suite of instruments and their characteristics of the platform used in
this study can be found in [Riley et al.,
2014](https://doi.org/10.1016/j.atmosenv.2014.09.018), [Larson et al.,
2017](https://doi.org/10.1016/j.atmosenv.2016.12.037), and [Xiang et
al.,2020](https://doi.org/10.1021/acs.est.0c00612). The remainder of the code in
this week's lab uses the GPS, P-Trak, and CO<sub>2</sub> measurements.


# Dates and Times

Managing dates and times in exposure assessment is a common challenge because
there are a variety of common date and time formats used to log instrument data.
You've no doubt encountered the following representations of dates:

   * `r format(Sys.time(), '%d %b, %Y')`
   * `r format(Sys.time(), '%m/%d/%Y')`
   * `r format(Sys.time(), '%Y-%m-%d')`

With direct-reading instruments made all over the world, you'll have the
potential to encounter many devices that use different date and time formats.
You'll find it advantageous to convert everything to a standardized format,
including the time zone. Consistently using a standardized date and time format
will make it less likely that you'll interpret a date and time incorrectly.

Date objects in `R` are of class `Date`, and compound date and time ("datetime")
objects, are of class `POSIXct` (shown as `dttm` when a tibble is printed to the
console). Let's take a look at  these object types and show their classes: 

```{r datetime.and.date.objects}
#-----datetime.and.date objects-----

# show POSIXct object -- a datetime object with both date and time
Sys.time()

# show class
# Note that 'POSIXt' is a virtual class which cannot be used directly. This
# virtual class exists from which both of the classes (POSIXct and POSIXlt)
# inherit. POSIXt is used to allow operations, such as subtraction, to mix the
# two classes.
class(Sys.time())

# show Date object-- date without time included
Sys.Date()

# show class
class(Sys.Date())

```

Internally, date and datetime objects store the number of seconds (for
`POSIXct`) or days (for `Date`) since January 1st 1970 at midnight. So for the
current time and date these values are `r as.numeric(Sys.time())` seconds and 
`r as.numeric(Sys.Date())` days. When printed to the console, they are displayed
in a human-readable format along the the International Organization for
Standardization (ISO) 8601 date and time standard. ISO 8601 provides an
unambiguous, well-known, and well-defined method of representing dates and
times.

It is often necessary to define date/time components or the format of a datetime
object specifically for each instrument. There are several reasons in practice
why we need to pay careful attention to this:  

  * the time scale of each instrument can be different, 
  * the internal instrument clocks can be different from each other and/or the
    current time, or
  * the instruments may be using different time zones, daylight savings time
    conventions, or there may be issues with simultaneous measurements on either
    side of the international date line.

Regardless, it is necessary to ensure a common schedule prior to merging the
data. Accounting for these issues during the initial data wrangling will make it
easier to handle date and datetime variables throughout your analysis process.


# Identify Files of Interest, Read Files & Wrangle

Here, we'll focus on CO<sub>2</sub> data from Car 1 (there were 2 cars), for your
assignment we ask you to analyze data from P-Traks, either using the same car or
the other car.

In each of the following chunks we show the tibble. In the `.html` file you can
scroll through the first 10,000 rows.  However, don't assume that 10,000 is the
maximum number of observations in the dataset. To learn that use `glimpse` or
`str`.

```{r gather.file.names.for.import}
#-----gather file names for import-----

# get filename paths for data files
filenames <- list.files(data_dir) %>% str_subset("car1")

```

## GPS

```{r gps, message=FALSE}
#-----gps-----

# Get GPS file name  
gps_file <- str_subset(filenames, "GPS")

# read GPS data
gps <- read_csv(file.path(data_dir, gps_file)) %>%
   
   # set column names
   set_names(c("date", "time", "latitude", "longitude", "altitude", "speed")) %>% 
   
   # create datetime column (notice we must specify the timezone)
   mutate(datetime = as.POSIXct(paste(date, time), tz = "US/Pacific") )

# show dataframe
gps

```

## P-Trak

These instruments collect ultrafine particles (UFPs) at 1-second time
resolution. An unscreened P-Trak collects particles in the range of 20 - 1,000
nm while the screened P-Trak collects particles in the range of 50-1,000 nm.
Thus, having both allows some rudimentary understanding of particle size
distributions. The instruments measure particle number concentration, with units
expressed as particles/cm^3^, #/cm^3^, pt/cm^3^, or 1/cm^3^.

```{r p.trak, message=FALSE}
#-----p-trak-----

# get p-trak not-screened file names
ptrak_noscreen_file <- filenames %>% 
   str_subset("PT") %>% 
   str_subset("scrnd|Scrnd|Screened|screen", negate = TRUE)

# read data
ptrak_noscreen <-  read_tsv(file.path(data_dir, ptrak_noscreen_file), skip = 29) %>% 
   
   # name columns
   set_names(c("date","time", "ptrak_noscreen_conc")) %>% 
   
   # modify date column and create datetime column (notice we must specify the
   # date format)
   mutate(date = as.Date(date, format = "%m/%d/%Y"), 
          datetime = as.POSIXct(paste(date, time), tz = "US/Pacific") )


# get p-trak-screened file names and read in data
ptrak_screen_file <- filenames %>% 
   str_subset("scrnd|Scrnd|Screened|screen")

# read data
ptrak_screen <-  read_tsv(file.path(data_dir, ptrak_screen_file), skip = 29) %>% 
   
   # name columns
   set_names(c("date","time", "ptrak_screen_conc")) %>% 
   
   # modify date column and create datetime column
   mutate(date = as.Date(date, format = "%m/%d/%Y"), 
          datetime = as.POSIXct(paste(date, time), tz = "US/Pacific") )


# show dataframes
ptrak_noscreen
ptrak_screen

```

## Carbon Dioxide

```{r co2, message=FALSE, warning=FALSE}
#-----co2-----

# get CO2 file name
co2_file <- str_subset(filenames, "CO2")

# read data, specify time as character because of time parsing issue
co2 <- read_tsv(file.path(data_dir, co2_file), skip = 1, 
                col_types = cols(`System_Time_(h:m:s)` = col_character()) ) %>% 
   
   # name columns
   setNames(c("date", "time", "co2_conc", "h2o_conc", "temp", "pressure", 
              "co2_absorp", "flow") ) %>% 
   
   # select columns of interest
   select(date, time, co2_conc) %>% 
   
   # address time parsing (some minute values have only one digit) and 
   # create datetime column
   mutate(time = as_hms(time), 
          datetime = as.POSIXct(paste(date, time), tz = "US/Pacific") )

# show dataframe
co2

```


# Merge & Inspect Instrument Data

The `purrr` package from `tidyverse` is a functional programming tool that
allows you to apply a function iteratively over a list.  We use that next to
merge (or "join") the datasets.  Note that the `full_join` function searches for
common column names and joins on these. Alternatively you can add this argument:
`by = c("date", "time", "datetime")`.

We print out the dataset to get a basic idea of whether the merge worked.  Ask
yourself questions like the following:

  * Are there missing data and what is its pattern?
  * Do the variables (columns) in the final dataset make sense to you?  

We observe that early rows of the dataset have missing GPS and CO<sub>2</sub>
data, so we know this is something we need to investigate further.

```{r merge.datasets}
#-----merge datasets-----

# create a list of all data to merge 
instrument_list <- list(gps = gps, ptrak_noscreen = ptrak_noscreen, 
                        ptrak_screen = ptrak_screen, co2 = co2)

# use `purrr::reduce` to run `full_join()` multiple times on `instrument_list`
all_data <- reduce(instrument_list, full_join) %>% 
   
   # reorder columns so datetime is first
   relocate(datetime) %>% 
   
   # arrange data in datetime order
   arrange(datetime)

# use glimpse to learn more about the data
glimpse(all_data)

```

## Plot Data 

Let's make a quick visualization of the concentration data to get an idea of the
data we're working with.

What features do you see?

```{r plot.data, warning=FALSE}
#-----plot data-----

# transform data from wide to long for ggplot
all_data %>% pivot_longer(cols = contains("_conc"), 
                          names_to = "pollutant", 
                          values_to = "concentration") %>% 
# pipe into ggplot
ggplot(aes(x = datetime, y = concentration)) + 
  
  # plot each pollutant in a facet
  facet_wrap(~pollutant, ncol = 1, scales = "free_y") + 
  
  # specify type of plot
  geom_line() +
  
  # specify theme
  theme_bw()

```

Data availability plots can be a useful initial check to visualize what data we
have across instruments over time. We show one variable from each dataset. In
this case the plot is not very informative because the data are fairly complete
and the plot doesn't make it easy to see the limited amount of missing data that
are present even though we already observed that there are missing data after
merging.

```{r data.availability.plot}
#-----data.availability.plot-----

# transform data from wide to long for ggplot (using latitude for GPS)
all_data %>% 
  mutate(GPS = latitude) %>%
  pivot_longer(cols = c(contains("conc"), GPS), 
               names_to = "measurement") %>%

# make plot
ggplot(aes(x = datetime, y = measurement, color = factor(measurement)) ) +
  geom_point(size = 0.5) + 
  labs(x = "Time", y = "Instrument") + 
  theme_bw() + 
  theme(legend.position = "none")

```

## Missing Data

Next we try to precisely determine how much missing data we have and what are
the likely sources of missingness in our data. First we calculate the total and
percent missing for every variable in the dataset. We observe that the presence
of missing data depends upon the instrument.

```{r missing}
#-----missing-----

lapply(all_data, function(i){ 
   
   tibble( 
          # sum missing
          n_miss = sum(is.na(i)), 
          
          # percent missing
          perc_miss = round(n_miss/length(i) * 100, 1)
          )
   }) %>% 
   
   # bind list
   bind_rows(.id = "variable")

```


# Time Interval

Our data appeared to be on a similar 1-second time interval. Let's confirm by
calculating the difference in time between consecutive measurements in each of
our datasets:

```{r time.stamps}
#-----time stamps

# we'll use the instrument data list
lapply(instrument_list, function(i){
   
   # get vector of datetimes
   datetimes <- i[["datetime"]]
   
   # get a vector of the differences between consecutive measurements
   time_diff <- datetimes[2:length(datetimes)] - datetimes[1:length(datetimes)-1]
   
   # summarize differences
   table(time_diff)
   
}) %>% 
   
   # use list names
   set_names(names(instrument_list))

```

It seems that some of our missingness is due to instrument reporting. Some
measurements did not occur at 1-second intervals. From this we can also see
there are duplicate measurements for some instruments (difference of zero
seconds).

Lastly, it appears that the GPS instrument was not turned on at the same time as
the other instruments (there is no GPS data at the beginning of the `all_data`
dataframe). This may be due to the fact that the other instruments were turned
on for "warm-up" before the measurement campaign started.

Let's drop data missing the GPS variables for convenience, but is this the best
thing to under all circumstances and for all variables?

```{r drop.missing.gps}
#-----drop missing gps-----

all_data <- drop_na(all_data, latitude, longitude)

```


# Timeseries Data

Time series data are marked by measurements that are indexed to a time component.
There are many `R` standards for time series data: `ts`, `xts`, `data.frame`,
`data.table`, `tibble`, `zoo`, `tsibble`, `tibbletime` or `timeSeries`. The
package `tsbox` has many useful functions for converting between these
time series formats.

We've focused most of our effort on `tidyverse` tools this term, so let's
concentrate on the `tsibble`, `feasts`, and `slider` package functions.

First we want to turn our dataset into a `tsibble`. We use the function `try` so
that knitting our document doesn't fail if this operation fails. We observe that
it does fail because there were duplicates, so then we inspect them.

```{r try.tsibble}
#-----try tsibble----

# use "try" here so document knits
try( as_tsibble(all_data, index = datetime) )

# inspect duplicate rows
duplicates(all_data, index = datetime)

```

`tsibble` alerts us of our time issues and prompts us to deal with them, so
let's remove the duplicates (you could average duplicates also). 

```{r to.tsibble}
#-----to tsibble-----

# remove duplicate rows, and convert to `tsibble`
ts_data <- all_data %>% 
  distinct(datetime, .keep_all = TRUE) %>% 
  as_tsibble(index = datetime)

```

Let's also fill the gaps in time. In other words, we have decided in this
analysis that the index column `datetime` needs to be unique and complete. In
standard time series analyses the time increment between every consecutive
observation is the same. Adding these new time rows introduces "explicit" NAs to
the data, so you'll then have a choice about if and how to fill in the non-time
data. `tidyr::fill()` can help with this task, and the choice of how to fill may
be more or less important depending on the time scale you're working on, the
amount of missingness, and if there are consecutive measurements missing. Given
the small amount of missing and the one-second time scale of the data, it's not
unreasonable to introduce a few rows with NAs here.

```{r fill.ts}
#-----fill ts-----

# count rows before filling
paste("number of rows before filling gaps:", nrow(ts_data))

ts_data <- fill_gaps(ts_data) 
  
# # if you wanted to "fill" the new explicitly missing data with a value drawn
# # from the dataset, here is an option. (add a `%>%` to the function above). 
# # Note this could have downstream effects. 
# fill(co2_conc, .direction = "down")

# count rows after filling
paste("number of rows after filling gaps:", nrow(ts_data))

# glimpse
glimpse(ts_data)

```

# Temporal Autocorrelation

Time series data is often correlated in time - measurements taken close in time
to one another are more similar than measurements taken farther apart. An
autocorrelation plot helps identify at what lag (or time interval) data are less
correlated. In the following plot, the blue lines indicate bounds on the
autocorrelation of these data, i.e., which autocorrelations statistically
different from zero. At what time do the CO<sub>2</sub> appear less correlated?

```{r autocorrelation}
#-----autocorrelation-----

ts_data %>% ACF(co2_conc, lag_max = 60) %>% autoplot()

```


# Temporal Smoothing with New Time Scales -- Moving averages

A common task with time series data is averaging to different time scales. From
the autocorrelation plot above, we can see at a lag of 60 seconds, the
CO<sub>2</sub> concentrations are no longer correlated. So let's convert our
1-second data to some longer time scales. One package that helps with averaging
tasks is `slider`.

For the first average we'll use the time scale period (minutes) to calculate
1-minute averages. Though with other data, this could also be months, weeks,
years, etc. This is a "block average" where the period we are averaging over is
based on an attribute of the data. For example, if you wanted monthly averages,
this method would be "aware" of the fact there are a different number of days in
each month. For the second average (a 2-minute average) we'll demonstrate a
moving ("sliding" or "rolling") average, where the number of neighbors before
and after a value of interest are used to compute an average.

Notice how both functions preserve the input dataframe size. For the period
average we can then easily merge the new column into our existing dataframe
(however, you'll notice values are repeated 60 times for each minute). The
moving average works nicely with `mutate()`.

## `slider`

```{r slider}
#-----slider-----

ts_data <- ts_data %>% 
  
  # left join new 1-minute period values
  left_join(

    # Approach 1:
    # use `slider` to calculate 1-minute concentrations by period
    slide_period_dfr(ts_data, ts_data$datetime, "minute",
                     ~tibble(datetime = floor_date(.x$datetime),
                             co2_conc_one = mean(.x$co2_conc)
                             ) ),
    by = "datetime" ) %>% 
  
  # add a datetime column with 1-minute precision to help if you choose to
  # `group_by()` and `summarise()` 1-minute averages
  mutate(datetime_one = ymd_hm(format(datetime, "%Y-%m-%d %H:%M"), 
                               tz = "US/Pacific")) %>%
  
  # Approach 2:
  # use `slider` to calculate 2 minute moving averages (this works because we've
  # made sure our datetime index is unique and complete)
  mutate(co2_conc_two = slide_dbl(co2_conc, ~mean(.x, na.rm = TRUE), 
                             .before = 60, .after = 60) ) 

# # slide_period also works in the same way, but it returns a vector of different
# # length than the original inputs
# with(ts_data, 
#      slide_period_dbl(.x = co2_conc, .i = datetime, 
#              .period = "minute", .f = mean, na.rm = TRUE) 
#      )

# glimpse
glimpse(ts_data)

```

## `tsibble`

Next, let's take a look at a `tsibble` approach for 5-minute averages. Notice
that the size of the dataframe decreases.

<!--TODO Chris:  Need to add 2.5 minutes to the datetime variable so this isn't offset in time -->

```{r new.timescale}
#-----new timescales-----

ts_new <- ts_data %>% 
  
  # get the "floor" of each datetime row (unfortunately `tsibble` doesn't let us
  # use "datetime" for this new variable name)
  index_by(datetime_new = floor_date(datetime, unit = "5mins")) %>%
  
  # summarise the mean of rows across all dataframe columns
  summarise(across(where(is.numeric), mean, na.rm = TRUE ), .groups = "drop") %>% 
  
  # rename to get "datetime" variable name back
  rename(datetime = datetime_new) %>% 
  
  # create new variable so new averages align with original due to `floor_date()`
  mutate(plot_time = datetime + minutes(2) + seconds(30))
  
# glimpse
glimpse(ts_new)

```

Following up on our autocorrelation plot above, notice how it changes. The lags
are now 5 minutes long and the evidence in the autocorrelation is only in the
first 2-3 lags. This is a longer duration in time than we saw with the 1-second
data, but a fewer number of lags.

```{r 5.min.autocorrelation}
#---- 5-min autocorrelation ----

# check 5-minute average autocorrelation
ts_new %>% ACF(co2_conc) %>% autoplot()

```

All our example data are on the 1-second time scale; however, it is common to
have multiple time scales in a suite of direct-reading instruments. Using the
techniques above, you'll be able to average different time intervals to a common
schedule. Once the measurements are on a shared time scale, they can be
merged/joined as we have shown in the "Merge & Inspect Instrument Data" section.

## Plot Different time scales

Let's inspect the time series of our different averages:

```{r plot.different.time.averages}
#-----plot different time averages-----

# we're going to use a simpler approach here and just manually specify each
# series rather than make a dataframe with all the plotting data combined

ggplot() + 
  
  # plot original 1-second data
  geom_line(data = ts_data, aes(x = datetime, y = co2_conc, color = "blue")) + 
  
  # plot `slider` 1-min period averages
  geom_line(data = ts_data, aes(x = datetime, y = co2_conc_one, color = "black")) +
  
  # plot `slider` 2-min moving averages
  geom_line(data = ts_data, aes(x = datetime, y = co2_conc_two, color = "red")) +
  
  # plot `tsibble` 5-minute block averages
  geom_line(data = ts_new, aes(x = plot_time, y = co2_conc, color = "darkgreen")) +  
  
  # specify legend values manually
  scale_color_identity(name = "Time Average", 
                       breaks = c("blue", "black", "red", "darkgreen"), 
                       labels = c("1-sec", "1-min", "2-min", "5-min"), 
                       guide = "legend") +
  
  # labels
  labs(x = "Time", 
       y = "CO2 (ppm)" 
       ) +
  
  # theme
  theme_bw() + 
  theme(legend.position = "bottom")

```


# Map concentrations

Our data was from a mobile monitoring campaign - so let's map the
concentrations.

Here is a helper function for the `bbox` argument of `get_stamenmap()`. It takes
a dataframe, finds the latitude and longitude columns, makes calculations to add
a little "extra" space to the plotting area, and outputs an object used to
define the extent of the mapping region. You might find it helpful in the
future. These operations require `stringr`and `tidyr`.

```{r define.make_bbox.function}
#-----define make_bbox function------

make_bbox <- function(.data, width_extra = 0.0001, height_extra = 0.00005){
  
  # dependencies
  pacman::p_load(stringr, tidyr)
  
  # find lat and long column names
  long <- str_subset(string = names(.data), pattern = "Long|long")
  lat <- str_subset(string = names(.data), pattern = "Lat|lat")
  
  # drop rows with NA
  temp <- .data %>% drop_na(all_of(c(long, lat)))
  
  # create a named vector with the corners of the plotting region
  out <- c(left = min(temp[[long]]) + width_extra*min(temp[[long]]),
           bottom = min(temp[[lat]]) - height_extra*min(temp[[lat]]),
           right = max(temp[[long]]) - width_extra*min(temp[[long]]),
           top = max(temp[[lat]]) + height_extra*min(temp[[lat]]) )
}

```

In the following chunk we define the bounding box (extent of the mapping area),
download the stamen map tiles (if they have not already been downloaded), and
save the map for later. If the map file has already been downloaded and saved,
we can just load the `map.rds` file. This will save you some time and from
unnecessarily repeating file downloads as you work to develop your code. Note if
you've already downloaded and saved a `map.rds` file, but then you want to
modify the underlying map somehow, you'll need to delete the `map.rds` file
first (or work interactively within the `if()` statement).

```{r get.map.objects}
#-----get map objects-----

# define the bounding box for the map using `make_bbox()` 
bbox <- make_bbox(ts_data)

# download stamen map if needed, otherwise just load it from a previous download
if (!file.exists(file.path(data_dir, "map.rds"))) {
  
  # make a map of the base layer of stamen tiles (without showing messages)
  map <- suppressMessages(get_stamenmap(bbox, zoom = 14, maptype = "terrain"))
  
  # save map to the mobile_monitoring data directory so you don't need to
  # download it again
  saveRDS(map, file = file.path(data_dir, "map.rds") )
  
  }

# read map file
map <- readRDS(file.path(data_dir, "map.rds"))

# make the map image from the stamen map tiles
basemap <- ggmap(map, darken = c(0.5, "white")) + theme_void()

```

To help represent the time-varying nature of the data in this lab, let's collect
the times of some of our measurements to help our visualization.

```{r get time stamps}
#-----get time stamps-----

map_times <- ts_data %>% 
  
  # convert to `tibble` because we're breaking up our `tsibble`
  as_tibble() %>%
  
  # filter to rows with :00, :15, :30, and :45 minute times
  filter(str_detect(time, pattern = ":00:00|:15:00|:30:00|:45:00")) %>% 
  
  # make a map_time variable, dropping the seconds 
  # (`$` indicates the character pattern should be at the end of the string)
  mutate(map_time = str_remove(time, pattern = ":00$"))

```

## Map Carbon Dioxide Concentrations

```{r map.concentration, warning=FALSE}
#-----map concentrations----

# map concentrations
map_co2 <- basemap +

  # locations with concentrations
  geom_point(data = ts_data %>% drop_na(co2_conc), 
             aes(x = longitude, y = latitude, color = co2_conc))+ 
  
  # add measurement times to map
  geom_text_repel(data = map_times,
                  aes(x = longitude, y = latitude, label = map_time), 
                  force_pull = -.02 ) +
  
  # color scale, chosen so that high values are darker
  scale_color_continuous(low = "#56B1F7", high = "#132B43") +  
  
  # labels
  labs(color = "CO2 (ppm)") +
  
  # theme
  theme_void() 

# show map
map_co2

```


# Background Standardization 

We show different approaches to estimating the background in this section and
plot them for comparison.

## Rolling Percentile & Rolling Minimum

[Brantley et al., 2014](https://amt.copernicus.org/articles/7/2169/2014/)
summarize a number of methods to calculate background concentrations. One
approach described by 
[Bukowiecki et al.,](https://doi.org/10.1016/S1352-2310(02)00694-5), includes 
taking the 5^th^ percentile of 1 or 5 min averages. As we talked about in class,
another approach is to calculate the moving minimum of 30-minute time intervals.
`slider` simplifies these types of moving calculations.

```{r calculate background.with.moving.functions}
#-----calculate background with moving functions-----

# specify the time in minutes for moving calculations 
# (for percentile `_pct` and minimum `_min`)
time_pct <- 5
time_min <- 30

# specify the percentile considered to be background
pct <- 0.05

# use use `slider` to add moving background calculations to `ts_data`. Then
# calculate an adjusted concentration by removing the background concentration.
# For moving percentile, assign new names with suffix `_pct`, and for moving
# minimums use `_min`. `min()` will return Inf when all values are NA; use
# `na_if()` to avoid.
ts_data <- ts_data %>% 
  mutate( 
         # using moving percentiles
         co2_background_pct = slide_dbl(co2_conc, 
                                    ~quantile(.x, probs = pct, na.rm = TRUE), 
                                    .before = (time_pct*60)/2, 
                                    .after = (time_pct*60)/2), 
         co2_adj_pct = co2_conc - co2_background_pct, 
         
         # using moving minimums
         co2_background_min = slide_dbl(co2_conc, 
                                 ~min(.x, na.rm = TRUE), 
                                 .before = (time_min*60)/2, 
                                 .after = (time_min*60)/2), 
         co2_background_min = na_if(co2_background_min, Inf), 
         co2_adj_min = co2_conc - co2_background_min
         )

```

For both the percentile approach we've taken `r time_pct` minutes
as the time interval and `r pct*100`^th^ percentile of the data to calculate the
background CO<sub>2</sub> concentration.

Let's compare our two background estimation approaches in one plot:

```{r plot.background, warning=FALSE}
#-----plot background-----

# plot the two background estimates on the same plot
ggplot(data = ts_data) + 
  geom_line(aes(x = datetime, y = co2_background_pct, color = "black")) + 
  geom_line(aes(x = datetime, y = co2_background_min, color = "red")) + 
  labs(x = "Time", y = "CO2 Background (ppm)") +

# specify legend values manually
scale_color_identity(name = "Background Type",
                   breaks = c("black", "red"),
                   labels = c("Percentile", "Minimum"),
                   guide = "legend") +
# theme
theme_bw() + 
theme(legend.position = "bottom")
  
```

## Loess Smoother of Background

We can further address the background by smoothing our estimates of background.
These background estimates, even when smoothed, appear to be picking up more
than background. However, in a later plot we consider the background estimates
in the context of the data, which are **much** higher, so the apparent high
variation in background isn't that large in the context of the complete
CO<sub>2</sub> dataset.


```{r loess.smooth.of.moving.background}
#-----loess smooth of moving background-----

# specify loess models
mdl_pct <- loess(co2_background_pct ~ seq_along(time), data = ts_data, 
                 span = 0.1, degree = 1)

mdl_min <- loess(co2_background_min ~ seq_along(time), data = ts_data, 
                 span = 0.1, degree = 1)

# add smoothed predictions of background to ts_data
ts_data <- ts_data %>% 
  mutate(co2_background_pctl = predict(mdl_pct, ts_data),
         co2_background_minl = predict(mdl_min, ts_data)) 

# prepare dataframe for ggplot 
# (creating new columns for our different types of estimates)
ts_data %>% pivot_longer(contains("_background_")) %>% 
  mutate(min_pct = if_else(str_detect(name, "min"), "minimum", "percentile"), 
         smooth_un = if_else(str_detect(name, "l"), "smoothed", "unsmoothed") ) %>% 

# compare all background estimates
ggplot(aes(x = datetime, y = value, 
           linetype = interaction(smooth_un, min_pct, sep = " "), 
           color = interaction(smooth_un, min_pct, sep = " ") ) ) + 
  
  # plot lines
  geom_line() + 
  
  # set colors
  scale_color_manual(name = "Background Type", 
                     values = c("red", "red", "black", "black")) +
  
  # set linetype
  scale_linetype_manual(name = "Background Type", 
                        values = c("dashed", "solid", "dashed", "solid")) + 
  
  # labels
  labs(x = "Time", y = "CO2 Background (ppm)") +
  
  # theme
  theme_bw() + 
  theme(legend.position = "bottom")

```

## Plot Carbon Dioxide with Background

We'll demonstrate with the the rolling minimum approach, showing both the
unsmoothed and loess smoothed versions.

```{r plot.co2.and.background, warning=FALSE}
#-----plot co2 and background-----

ggplot(data = ts_data) + 
  
  # plot the co2 observations
  geom_point(aes(x = datetime, y = co2_conc), size = 0.8, color = "darkgray") +
  
  # plot co2 background using the loess smoothed and unsmoothed rolling minimums
  geom_line(aes(x = datetime, y = co2_background_minl), lty = "dashed", color = "red") + 
  geom_line(aes(x = datetime, y = co2_background_min), lty = "solid", color = "red") + 
  
  # labels
  labs(x = "Time", y = "CO2 with Background (ppm)") +
  
  # theme
  theme_bw()

```


# Identifying Peak Concentrations

<!-- Elena, Please add to or edit as needed, this was material I wasn't so sure
of. In particular, whether or not  -->

As we've seen, mobile monitoring data has a mix of temporal and spatial
variability. In an effort to disentangle temporal confounding from spatial
analysis, we can calculate the background concentration of a pollutant. Peak
identification can then be performed on the background adjusted data to identify
single or multi-pollutant extreme features (for example, departures above a
particular quantile of single pollutant features vs departure above background
of a multivariate representation such as a PCA feature).

With background removed from the CO<sub>2</sub> concentrations, we can now use
the adjusted data to identify concentration peaks. Concentration peaks indicate
potential sources of the pollutant. In this case, a CO<sub>2</sub> peak could be
a large vehicle like a truck.

```{r id.peaks, warning=FALSE}
#-----id peaks-----

# specify the time in minutes from which to calculate the percentile
time_pct <- 5

# specify the percentile considered to be peaks
pct <- 0.95

# we can use use `slider` to calculate these rolling statistics
# use our previous adjusted co2 from the rolling 5th percentile (co2_adj_pct)
ts_data <- ts_data %>% 
  mutate(co2_peak = slide_dbl(co2_adj_pct, 
                              ~quantile(.x, probs = pct, na.rm = TRUE), 
                              .before = (time_pct*60)/2, 
                              .after = (time_pct*60)/2), 
         co2_is_peak = if_else(co2_adj_pct >= co2_peak, TRUE, FALSE)
         ) 

```

```{r plot.co2.peaks, warning=FALSE}
#-----plot co2 peaks-----

ggplot(ts_data) + 
  
  # plot co2 observations
  geom_point(aes(x = datetime, y = co2_conc, color = co2_is_peak), size = 0.8) +
  
  # plot background
  geom_line(aes(x = datetime, y = co2_background_pct)) + 
  
  # set colors
  scale_colour_manual(values = c("darkgray", "darkred")) +
  
  # labels
  labs(x = "Time", y = "CO2 Peaks with Background (ppm)") +
  
  # theme
  theme_bw() + 
  theme(legend.position = "none")

```


# Local Emission Plume Detection with P-Trak 

<!-- Elena, 
THIS WAS IN KOLB (REFERENCED IN BRANTLEY) AND SEEMED DOABLE, USING THE IDEA OF
RATIOS AND POTENTIALLY INTERESTING. IT TURNED OUT TO BE KIND OF BORING (OR I'VE
MESSED SOMETHING UP)...

IF YOU'D LIKE TO KEEP THE TEXT LIKELY COULD USE SOME EDITING/ADDITIONS -->

As the CO<sub>2</sub> map shows, the data we've collected represent
concentrations over 4 hours, meaning some of the variability we're observing
could be a mix of spatial (i.e. the variability we see in the map) and temporal
(i.e. the variability observed in time series plots) processes.

Next, we'll use the P-Trak emission ratio data to explore the idea of pollutant
emission ratios described in ([Kolb et al.,
2004](https://doi.org/10.1021/es030718p)), which normalizes pollutant
concentrations to CO<sub>2</sub> concentrations, potentially helping identify
emission sources in data with both temporal and spatial variability.

```{r map.emission.ratios, warning=FALSE}
#-----map emission.ratios----

# calculate ratio, this time using adjusted co2 from the rolling minimum
# (co2_adj_min)
ts_data <- ts_data %>% 
  mutate(ptrak_ratio = ptrak_screen_conc / co2_adj_min)

# map concentrations
map_ratios <- basemap +
    
# locations with ratios
geom_jitter(data = ts_data %>% drop_na(ptrak_ratio), 
           aes(x = longitude, y = latitude, color = ptrak_ratio)) + 

# color scale
scale_color_continuous(low = "#56B1F7", high = "#132B43") +  

# labels
labs(color = "P-trak (screened) \nEmission Ratio") +

# theme
theme_void() 

# show map
map_ratios

```


# Homework

Use the data from car1 or car2 to repeat a subset of the analyses demonstrated
above for ultrafine particles and write up a coherent report about the patterns
you see. The goal of this lab is for you to use the code we have provided to
read in time-varying data from multiple instruments, check that they were read
in appropriately, and address a few scientific questions about time-varying
exposures. Here are the topics for you to address in your lab report:

1.  Data description, data quality, and approach to analysis:
    a.  How many observations were available, and how many were usable after 
        preliminary data management (e.g. dropping NAs after merging)?
    b.  What variables are available and which ones are of scientific interest?
    c.  When were the data collected and in what area?
    d.  What steps did you follow to accomplish the scientific analyses you 
        discuss in the rest of the report?

2.  Scientific analyses:
    a.  Plot a time series plot for the screened and unscreened P-Trak data on 
        one plot.  Do you see a different pattern for the two measurements of 
        ultrafine particles?
    b.  Look at the autocorrelation in the unscreend P-Trak data. Compare 
        autocorrelation plots at the 1-second time scale with autocorrelation 
        plots at the 5-minute time scale. Describe the patterns and comment on 
        their similarities and differences.
    c.  Choose one additional analysis of interest to you that will allow you to
        highlight a feature of the time-varying data that is of scientific 
        interest to you. Justify your reasoning for your choice, leveraging your
        understanding from the lecture and/or Brantley paper.  For instance, you
        might consider comparing different background estimation approaches, 
        showing a time series with background overlaid or removed, comparing the
        difference between the screened and unscreened P-Traks over space and/or
        time, or, identifying peak concentrations. Present an appropriate 
        choice of plots and/or maps to support your analysis. Interpret these 
        and connect your interpretation to the material discussed in the 
        lecture.


# Code Appendix

## Session Information

```{r session.info}
#-----session info: beginning of Code Appendix-----

sessionInfo()

```

## Code in the R Markdown file

```{r appendix.code, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60), include=T}
#-----appendix code-----

```

## User-written functions loaded in the R Markdown environment

```{r functions.used.in.this.Rmd, eval = TRUE}
#-----functions used in this Rmd-----

# Show the names of all functions used (loaded in the current environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lapply(c(lsf.str()), getAnywhere)

```

