---
title: "Week 5 Lab:  Variance Components"
author: "Lianne Sheppard for ENVH 556"
date: "Winter 2021; Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        fig_caption: yes
        toc: true
        toc_depth: 3
        number_sections: true
---

```{r setup, include=FALSE}
#-----setup-----

# set knitr options
knitr::opts_chunk$set(echo = TRUE)

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.
pacman::p_load(knitr, dplyr, tidyr, modelr, stringr, ggplot2, Hmisc, lme4, VCA)

```

```{r directory.organization.read.data, echo=FALSE, warning=FALSE}
#-----directory organization and read data-----

# specify working directory
project_path <- getwd()

# create "Datasets" directory if one does not already exist    
dir.create(file.path(project_path,"Datasets"), showWarnings=FALSE, recursive = TRUE)

# specify data path
data_path <- file.path(project_path,"Datasets")

# specify the file name and path
file_name <- "weldschool.rds"
file_path <- file.path(data_path, file_name)

# Download the file if it is not already present
if (!file.exists(file_path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 file_name, sep = '/')
    download.file(url = url, destfile = file_path)
}

# Output a warning message if the file cannot be found
if (file.exists(file_path)) {
    weldschool <- readRDS(file_path)
} else warning(paste("Can't find", file_name, "!"))

# remove temporary variables
rm(url, file_name, file_path, data_path)

```

# Introduction and Purpose

The purpose of this lab is to practice estimating variance components both by
using descriptive statistics as well as through application of random and mixed
effect models. In addition, as with all the labs in this course, you should
focus on writing up a coherent lab report that addresses the scientific context
and scientific questions that can be answered using variance components.

In this lab exercise we will use the Welding School personal survey data.  The
data are stored in the file named “*weldschool.rds*”.  The exposure
concentration variables in the dataset consist of *mnconc* and *particulate*,
the airborne concentrations of manganese and total particulate, respectively,
both in µg/m^3^.  The grouping variable is *subjid*.  Further description of the
variables can be found in the document with title "*Welding School Exposures variables 556
rev021015*", and the background and study design is in the published paper
with file name “*Baker Blood Mn Review*.”

# Getting Started

The companion file *Week5Lecture.Rmd* gives R commands relevant to this lab. The
lecture examples are applied to the DEMS underground and Snapshot traffic
gradient site data.  Unlike previous labs, we do not provide you with worked
examples from the Welding School dataset.  You will need to apply the tools
demonstrated in the *Week5Lecture.Rmd* to your analyses for this lab.

## Tools for estimating variance components

Below is a summary of R packages.  See the *Week5Lecture.Rmd* file for coding
examples.

* Descriptively estimate variance components using `dplyr`
* Random effects models only
    - ANOVA method of moments using `VCA`
    - ANOVA restricted maximum likelihood using `VCA`
* Random effects and mixed models
    - ANOVA restricted maximum likelihood using `lmer` in the `lme4` package
    - ANOVA maximum likelihood using `lmer` in the `lme4` package

## Overview of `lmer` commands

### `lmer` overview

The `lmer` function in the `lme4` package looks similar to a standard linear
regression model, but it adds capacity to incorporate structured errors as
random effects.  You can fit a variety of random effects models including the
simplest (a single random intercept), to more complex with multiple random
effects terms, nested (v.s. crossed) terms, and random slope models.

The formula part of the lmer command allows random effects to be specified with
terms inside parentheses.  The vertical bar separates the formula part of the
random effect model specification (before "|") from the variable that defines
the random effect (after "|") .

Fitting of `lmer` models is done by maximum likelihood estimation.  The random
effects and residual (unstructured) error in this model are assumed to have
normal distributions.  The default estimation uses restricted maximum
likelihood, which corrects for the bias in the variance component estimates.
However, when comparing models with different number of terms, standard maximum
likelihood should be used.

### `lmer` random effects coding

Here is a
[link](https://stats.stackexchange.com/questions/13166/rs-lmer-cheat-sheet) to a
helpful stackexchange post on `lme4` random effects coding.  Here is a brief
summary for categorical (grouping) variable G and continuous variable X where
our goal is to have random effects for levels of G:

* Random intercept only: (1|G)
* Random slope only:  (0 + X|G)
* Correlated random slopes and intercepts: (1 + X|G)
* Independent random slopes and intercepts: (1|G) + (0 + X|G)  



# Practice Session

1.  Determine your project for this lab.

2.	Read in the *weldschool.rds* dataset:
    a.	Become familiar with your data
    b.	How many observations, students, quarters are in your dataset?  How many
        observations are there per student (average, minimum, maximum)?
    c.	Characterize the outcome variables *mnconc* and *particulate*, airborne
        measurements of Mn concentration and total particulate, respectively, 
        and all potential covariates.
    d.	Get a basic understanding of the distribution of the airborne
        measurements on both the native and log-transformed scale.  What scale 
        do you think you should conduct your variance components analysis on?
    
For the rest of the practice session we will focus on the Mn concentration
variable:

3.	Try to estimate the variance components within and between subject "by hand"
    using `dplyr`.

4.	Now use `VCA` to estimate the variance components using a method of moments
    or ANOVA approach.
    a.	Do your estimates agree with those from Step 3?  Why or why not?

5.	Now use `lmer` to estimate the variance components.  (Decide whether you want
    maximum likelihood or restricted maximum likelihood estimates.  Be able to
    articulate your reason for your choice.)
    a.	Do your estimates agree with those from Step 3?  Why or why not?

6.	Decide whether you should add any fixed effects into your model.  
    a.	If so, which one(s)?  Why?  (Note:  Make sure to consider science in your
        answer!)
    b.	Compare the variance components from the models with and without fixed
        effect(s).  What happens to the variance component estimates from Step 5
        after you add your selected covariate(s)?  Does this make sense to you?
        (Hint:  Consider how your covariate is distributed within and between
        subjects in the data.)

# Homework Exercises

1.	Repeat the practice session analysis with particulate concentration.  Does
    this produce any different results? Why or why not?

2.	How do your variance component estimates compare with the DEMS and Snapshot
    examples we discussed in lecture?

3.	Report and discuss the results of your analysis of Mn and total particulate
    for this lab, touching on the following points:
    a.	The design of the study and structure of the dataset.
    b.	The scale of the data on which you did your analysis.
    c.	The comparability of variance component estimates using different
        estimation approaches and some possible reasons for (lack of) 
        comparability.
    d.	The impact on the variance component estimates when you include fixed
        effect(s) in your model and the reason for the change(s) in these data.
    e.	Incorporate insights from the [Peretz et al,
        2002](https://doi.org/10.1093/annhyg/mef009) paper.

# Appendix

```{r session.info}
#-----session information----

# print R session information
sessionInfo()

```

```{r code.appendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE, , include=TRUE}
#-----code appendix----
```

```{r functions, eval = TRUE}
#-----functions----

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lapply(c(lsf.str()), getAnywhere)

```
