---
title: 'Week 3 Lab:  Regression for Association'
author: "Lianne Sheppard for ENVH 556"
date: "Winter 2021; Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        df_print: "paged"
        toc: true
        toc_depth: 3
        number_sections: true
---

```{r setup, include=FALSE}
#-----setup-----

# set knitr options
knitr::opts_chunk$set(echo = TRUE)

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
   
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.
pacman::p_load(knitr, dplyr, ggplot2, egg, multcomp, modelr)

```

```{r directory.organization.read.data, echo=FALSE, warning=FALSE}
#-----directory organization and read data-----

# specify working directory
project_path <- getwd()

# create "Datasets" directory if one does not already exist    
dir.create(file.path(project_path,"Datasets"), showWarnings=FALSE, recursive = TRUE)

# specify data path
data_path <- file.path(project_path,"Datasets")

# specify the file name and path
file_name <- "allseasonsR.rds"
file_path <- file.path(data_path, file_name)

# Download the file if it is not already present
if (!file.exists(file_path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 file_name, sep = '/')
    download.file(url = url, destfile = file_path)
}

# Output a warning message if the file cannot be found
if (file.exists(file_path)) {
    snapshot<-readRDS(file_path)
} else warning(paste("Can't find", file_name, "!"))

# remove temporary variables
rm(url, file_name, file_path, data_path)

```

# Purpose

The purpose of this lab is to get practice using and reporting regression models with a focus on regression for association and some beginning work on regression for prediction.   We will use the snapshot data for this lab.

# Getting Started

This section gives some basic R commands for regression for association and the first steps for evaluating predictions.  

* Summarizing *ln_nox* by season  

```{r summarize.ln_nox}
#-----summarize ln_nox-----

# frequencies only
summary(snapshot$seasonfac)

# descriptive statistics
snapshot_summary <- snapshot %>%     
    group_by(seasonfac) %>%
    summarise(N = n(),
              Mean = mean(ln_nox), 
              SD = sd(ln_nox),
              GM = exp(mean(ln_nox)), 
              GSD = exp(sd(ln_nox)), 
              .groups = "drop")

# kable table
kable(snapshot_summary, digits=2)

```

* Commands for season-specific boxplots:
```{r season-specific boxplots}
#-----season-specific boxplots-----

ggplot(snapshot, aes(x = seasonfac, y = ln_nox, fill = seasonfac)) +
    geom_boxplot() +
    labs(x = "Season", 
         y = "ln(NOx) (ln(ppb)", 
         color = "Season", 
         fill = "Season") + 
    theme_article()
       
```

* Commands for regression: 

```{r regression for association}
#-----regression for association-----

# Common model (Table 4), season-specific LUR

# specify regression model
frml <- as.formula(ln_nox ~ D2A1 + A1_50 + A23_400 + Pop_5000 + D2C + Int_3000 + D2Comm)

# summarize the fit of each model
summary(lm_summer <- lm(frml, data = snapshot, subset = seasonfac == "1Summer"))
summary(lm_fall <- lm(frml, data = snapshot, subset = seasonfac == "2Fall"))
summary(lm_winter <- lm(frml, data = snapshot, subset = seasonfac == "3Winter"))

```

* Making predictions:

This gives season-specific predictions in the dataset with all seasons.

```{r predictions.with.dplyr}
#-----predictions with dplyr-----

# this adds model-specific predictions, both in-sample and out-of-sample
# however, we don't get the prediction intervals or SEs this way
snap2 <- snapshot %>% 
  
    # summer
    add_residuals(lm_summer,"resids_sum") %>%
    add_predictions(lm_summer,"preds_sum") %>%
    
    # fall
    add_residuals(lm_fall,"resids_fall") %>%
    add_predictions(lm_fall,"preds_fall") %>%
    
    # winter
    add_residuals(lm_winter,"resids_win") %>%
    add_predictions(lm_winter,"preds_win")

# inspect predictions and residuals
snap2 %>% 
  dplyr::select(SeasonID = seasonfac, ln_nox, preds_sum, resids_sum, preds_fall,
                resids_fall, preds_win, resids_win) %>% 
  mutate_if(is.double, round, 2)

```

Now we add prediction intervals:

```{r in-sample prediction intervals}
#-----in-sample prediction intervals-----

# first get the prediction interval for each season and bind it to the
# season-specific subset of the full dataset
# NOTE:  We are assuming the two datasets are in the same order when we cbind!
# It is good practice to check and not assume this is correct.
# R gives a warning about the interpretation of these prediction intervals when
# the data to generate the model are the same ones to do prediction.  This
# warning does not appear when the predictions are out of sample (see next
# chunk).
summer <- cbind(snapshot[snapshot$season==1,],predict(lm_summer,interval="prediction"))
fall <- cbind(snapshot[snapshot$season==2,],predict(lm_fall,interval="prediction"))
winter <- cbind(snapshot[snapshot$season==3,],predict(lm_winter,interval="prediction"))

# then combine the dataset into one big tibble
allseas_in <- rbind(summer, fall, winter) %>% as_tibble()

# now use dplyr to rename the predictions (the predict function gives default
# variable names)
allseas_in <- rename(allseas_in, pred_in = fit, lwr_in = lwr, upr_in = upr)

```

```{r out-of-sample predictions}
#-----out-of-sample predictions-----

# This example produces out-of-sample season-specific prediction intervals using
# the previous season to predict the next one.  (Note you should think carefully
# from a scientific perspective about how to approach out-of-sample predictions
# like these.)

# get the prediction interval for each season 
fall_preds_from_summer <- 
  predict(lm_summer, snapshot[snapshot$season == 2,], interval = "prediction")

winter_preds_from_fall <- 
  predict(lm_fall, snapshot[snapshot$season == 3,], interval = "prediction")

summer_preds_from_winter <- 
  predict(lm_winter, snapshot[snapshot$season == 1,], interval = "prediction")

# then combine the dataset into one big tibble and rename variables
allseas_out <- rbind(summer_preds_from_winter, 
                     fall_preds_from_summer,
                     winter_preds_from_fall) %>% 
  as_tibble() %>% 
  rename(pred_out = fit, lwr_out = lwr, upr_out = upr)

# Bind "out" predictions to the previous dataset with "in" sample predictions
# NOTE:  We assume the datasets are in the same order!
allseas_both <- cbind(allseas_in, allseas_out)

```

Now evaluate the quality of the predictions.  This is based on correlation between the prediction and the outcome.  It is the R^2^ for the best fit line of the relationship between the predictions and the data (*ln_nox* here).  It does not account for the systematic bias in the predictions.  Next week we will learn how to compute R^2^ about the 1:1 line which also addresses the systematic bias of the predictions.

```{r prediction assessment}
#-----prediction assessment-----

# summer in-sample
paste("summer in-sample R2:  ", 
      with(subset(allseas_both, season == 1), round(cor(ln_nox, pred_in)^2, 3))) 

# summer out-of-sample (from winter model)
paste("summer out-of-sample R2 (from winter model):  ", 
      with(subset(allseas_both, season == 1), round(cor(ln_nox, pred_out)^2,3)))

```

Let's look at what we mean by R^2^ about the best fit line by plotting the data we correlated above and incorporating the best fit lines.

```{r summer.in.and.out.sample, message=FALSE}
#-----summer in and out of sample-----

# plot the predictions vs. data for summer, both in-sample and out of sample

# get the range of the data to use as limits in the plots
r <- allseas_both %>% 
  filter(season == 1) %>% 
  dplyr::select(pred_in, pred_out, ln_nox) %>% 
  range()

# in-sample plot
p_in <- allseas_both %>%
    filter(season == 1) %>%
    ggplot(aes(x = pred_in, y = ln_nox)) +
    geom_point() +
    lims(x = c(r[1], r[2]), y = c(r[1], r[2]) ) +
    coord_fixed() +
    geom_abline(intercept = 0, slope = 1, color = "blue") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = "In-sample", 
         x = "In-sample predicted ln(NOx) (ln(ppb))",
         y = "Observed ln(NOx) (ln(ppb))"
         ) + 
    theme_bw()
 
# out-of-sample plot (predictions from the winter model)
p_out <- allseas_both %>%
    filter(season == 1) %>%
    ggplot(aes(x=pred_out, y=ln_nox)) +
    geom_point() +
    lims(x = c(r[1], r[2]), y = c(r[1], r[2]) ) +
    coord_fixed() +
    geom_abline(intercept = 0, slope = 1, color = "blue") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = "Out-of-sample (from winter model)", 
         x = "Out-of-sample predicted ln(NOx) (ln(ppb))",
         y = "Observed ln(NOx) (ln(ppb))"
         ) + 
    theme_bw()

# combine plots into one display
ggarrange(p_in, p_out, ncol = 2, 
          top = "Summer Model ln(NOx) Predictions", 
          bottom = "Best fit line is red; 1:1 line is blue")

```

# Practice Session
This section covers basic practice to be completed during the lab.   

Perform the following tasks: 

1.	Set up/decide your project for this lab.  
2.	Read in the snapshot data and get some basic understanding of its structure.  See the lecture `.Rmd` file for some options. 
3.	Summarize *ln_nox*, the outcome variable of interest     
    a.	Summarize by season    
    b.	Can you produce a nice summary figure of the data?  Should they be on the log or the native scale?  Is it useful to put multiple pollutants (NO, NOx, NO2) in the same figure?  
4.	Replicate the season-specific models in Table 4 of [Mercer et al, 2011](https://doi.org/10.1016/j.atmosenv.2011.05.043).  (Only focus on the LUR model results.  Also we’ll talk about cross-validation in an upcoming week so recognize your in-sample R^2^ should be a bit bigger and your RMSE should be a bit smaller than the values in the paper.)  
    a.	Compare your coefficient estimates and standard errors, as well as LUR R^2^ and sample size.  (Note:  You may need to consider rounding in your comparison.)  
    b.	Are all your terms parameterized the same as they are in the paper?  
5.	Use the fall model to predict ln_nox in summer and vice versa.    
    a.	Assess the quality of the predictions.  (Using the information from the lecture on regression for association, compute the R^2^, and plot the predictions and the prediction intervals.)  
    b.	What have you learned?  Does this flipping of models across seasons make sense scientifically?  Why or why not?  


# Homework Exercises  

1.	Describe the NOx variable.  Develop one figure and one table to best capture your data and then write a paragraph describing what these show.  (Doing both is for practice.  In a peer-reviewed paper you will ordinarily only show one of these.)  In the table, you may find it helpful to also include information on the covariates you use in your model(s) in this lab.    
2.	Using the terminology given in lecture, briefly discuss the fall season common LUR model results.  Include in your discussion an interpretation for at least two terms in the fall season model.   
3.	Focusing on the common models in Table 4 of [Mercer et al, 2011](https://doi.org/10.1016/j.atmosenv.2011.05.043), think about how to use the data from all seasons at once to get season-specific parameter estimates.   What terms do you need to incorporate the interactions?   
    a.  How is your interaction model different from fitting three separate season-specific models?
4.	Make table(s) and/or figure(s) summarizing your results for practice exercise 5.  Discuss.  

# Appendix

```{r session.info}
#-----session information----

# print R session information
sessionInfo()

```

```{r code.appendix, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE, , include=TRUE}
#-----code appendix----
```

```{r functions, eval = TRUE}
#-----functions----

# Show the names of all functions defined in the .Rmd
# (e.g. loaded in the environment)
lsf.str()

# Show the definitions of all functions loaded into the current environment  
lapply(c(lsf.str()), getAnywhere)

```


