---
title: "Week 9 Lecture Commands"
author: "Lianne Sheppard for ENVH 556"
date: "Winter 2021; Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        fig_caption: yes
        toc: true
        toc_depth: 3
        number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#-----setup-----

# set knitr options
knitr::opts_chunk$set(echo = TRUE)

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.
pacman::p_load(knitr, dplyr, feasts, slider, tidyr, ggplot2)
# may not need older time series packages:
pacman::p_load(tseries, xts) 

```

# Slides 8 & 9:  AR(1) process example, $\alpha$=0.5

```{r AR(1) example}
#---- AR(1) example----

# function to generate an AR(1) series.  Includes units in function
ar_data <- function(n = 500, interval = 1, units = "seconds", alpha=0.5){
  
  # function arguments
    # n: integer, desired output dataframe length
    # interval: integer specifying the number between timestamps
    # units: character specifying the lubridate period function (e.g. "minutes")
    # alpha: numeric value less than 1 specifying the AR alpha term.
  
  # function requires lubridate and tsibble
  pacman::p_load(lubridate, tsibble)
  
  # use match.fun() to find the lubridate function specifying time units
  input_units <- match.fun(units)
  
  # generate time sequence
  x <- seq(1:n)*interval %>% input_units()
  
  # generate e and y
  e <- rnorm(n)
  y <- double(n)
  y[1] <- 0
  for (i in 2:n) {y[i] <- alpha*y[i-1] + e[i]}

  # make a `tsibble`
  ts_ex_data <- data.frame(time = x, y) %>% as_tsibble(index = time) 
  
  # explicit return
  return(ts_ex_data)
  
}


# Alternative to function:  Manual creation of an AR(1) time series
# # n is length of time series
# n <- 500
# # generate data with alpha = .5
# e <- rnorm(n)
# y <- double(n)
# y[1] <- 0
# for (i in 2:n) {y[i] <- 0.5*y[i-1] + e[i]}

set.seed(20)

# create data 
ts_ex_data <- ar_data() 

# plot time series
ggplot() + 
  geom_line(data = ts_ex_data, aes(x = time, y = y)) +
  labs(title = "AR(1) time series",
       x = "Time axis",
       y = "Outcome")
 
# plot autocorrelation
ts_ex_data %>% ACF(y) %>% autoplot()

# plot partial autocorrelation
ts_ex_data %>% PACF(y) %>% autoplot()

```

# Slide10:  AR(1) process example, $\alpha$=0.9

```{r AR(1) example2}
#---- AR(1) example2----

# create data 
ts_ex_data <- ar_data(alpha = 0.9) 

# plot time series
ggplot() + 
  geom_line(data = ts_ex_data, aes(x = time, y = y)) +
  labs(title = "AR(1) time series, alpha=0.9",
       x = "Time axis",
       y = "Outcome")
 
# plot autocorrelation
ts_ex_data %>% ACF(y) %>% autoplot()

# plot partial autocorrelation
ts_ex_data %>% PACF(y) %>% autoplot()

```

# Slide 13:  Moving average example

```{r MA(2) process example}
#---- MA(2) process example----

# n is length of time series
n <- 500

# generate data with rho1 = .0, rho2 = .8
e <- rnorm(n)
y <- double(n)
y[1] <- e[1]
y[2] <- e[2] + 0.8*e[1]
for (i in 3:n) {y[i] <- e[i] + 0.8*e[i-1] + 0.8*e[i-2]}
 
# make a tsibble 
ts_ex_data <- tibble(y, time = seq(1,n)) %>%
  as_tsibble(index = time)

# plot time series
ggplot() + 
  geom_line(data = ts_ex_data, aes(x = time, y = y)) +
  labs(title = "MA(2) time series",
       x = "Time axis",
       y = "Outcome")
 
# plot autocorrelation
ts_ex_data %>% ACF(y) %>% autoplot()

# plot partial autocorrelation
ts_ex_data %>% PACF(y) %>% autoplot()



```

# Slide 14:  ARMA(1,1)

```{r ARMA(1,1)}
#---- ARMA(1,1) ----

set.seed(200)

# n is length of time series
n <- 500

# generate data with alpha=0.9, rho = .8
e <- rnorm(n)
y <- double(n)
y[1] <- e[1]
for (i in 2:n) {y[i] <- 0.9*y[i-1] + e[i] + 0.8*e[i-1] }
 
# make a tsibble 
ts_ex_data <- tibble(y, time = seq(1,n)) %>%
  as_tsibble(index = time)

# plot time series
ggplot() + 
  geom_line(data = ts_ex_data, aes(x = time, y = y)) +
  labs(title = "ARMA(1,1) time series",
       x = "Time axis",
       y = "Outcome")
 
# plot autocorrelation
ts_ex_data %>% ACF(y) %>% autoplot()

# plot partial autocorrelation
ts_ex_data %>% PACF(y) %>% autoplot()

```

# Not a slide:  Time plots with smoother overlaid

The next chunk shows the slider function and a user-written moving function that can present a running 10th percentile.

```{r Time plot with smoother}
#---- Time plot with smoother ----

# Using slider function with mean; the lab shows it with the quantile
ts_ex_data <- ts_ex_data %>%
  mutate(slide_20 = slide_dbl(ts_ex_data$y, ~mean(.x, na.rm = TRUE), .before = 10, .after = 10) )


# User-written moving window function:  Functionally this is the same as the
# slide function; seeing the code allows you to see how it can be constructed.
# Code found here:
# https://stackoverflow.com/questions/743812/calculating-moving-average
moving_fn <- function(x, w, fun, side, ...) {
  # x = vector with numeric data
  # w = window length
  # fun = function to apply
  # side = side to take, (c)entre, (l)eft or (r)ight
  # ... = parameters passed on to 'fun'
  y <- numeric(length(x))
  for (i in seq_len(length(x))) {
    if (side %in% c("c", "centre", "center")) {
      ind <- c((i - floor(w / 2)):(i + floor(w / 2)))
    } else if (side %in% c("l", "left")) {
      ind <- c((i - floor(w) + 1):i)
    } else if (side %in% c("r", "right")) {
      ind <- c(i:(i + floor(w) - 1))
    } else {
      stop("'side' must be one of 'centre', 'left', 'right'", call. = FALSE)
    }
    ind <- ind[ind %in% seq_len(length(x))]
    y[i] <- fun(x[ind], ...)
  }
  y
}

# and now create any variation you can think of!
moving_average <- function(x, w = 5, side = "centre", na.rm = FALSE) {
  moving_fn(x = x, w = w, fun = mean, side = side, na.rm = na.rm)
}

moving_sum <- function(x, w = 5, side = "centre", na.rm = FALSE) {
  moving_fn(x = x, w = w, fun = sum, side = side, na.rm = na.rm)
}

moving_maximum <- function(x, w = 5, side = "centre", na.rm = FALSE) {
  moving_fn(x = x, w = w, fun = max, side = side, na.rm = na.rm)
}

moving_median <- function(x, w = 5, side = "centre", na.rm = FALSE) {
  moving_fn(x = x, w = w, fun = median, side = side, na.rm = na.rm)
}

moving_Q1 <- function(x, w = 5, side = "centre", na.rm = FALSE) {
  moving_fn(x = x, w = w, fun = quantile, side = side, na.rm = na.rm, 0.25)
}

moving_Q3 <- function(x, w = 5, side = "centre", na.rm = FALSE) {
  moving_fn(x = x, w = w, fun = quantile, side = side, na.rm = na.rm, 0.75)
}

moving_10th <- function(x, w = 5, side = "centre", na.rm = FALSE) {
  moving_fn(x = x, w = w, fun = quantile, side = side, na.rm = na.rm, 0.10)
}

# Test rolling background:
roll10_y <- moving_10th(ts_ex_data$y, w = 10, side = "c")

# Show the results on a plot
ts_ex_data %>%
  bind_cols(roll10_y) %>%
  ggplot(aes(time, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(time, slide_20), color="red", lty=1) +
  geom_line(aes(time, roll10_y), color="blue", lty=2) +
  labs(title = "ARMA(1,1) time series\n slider & rolling 10th %tile",
       x = "Time axis",
       y = "Outcome")

```

# Loess and kernel smoother examples

Examples in this section follow the nice exposition by Rafael A. Irizarry in his
[Introduction to Data Science](https://rafalab.github.io/dsbook/) on [bin
smoothing](https://rafalab.github.io/dsbook/smoothing.html#bin-smoothing) and [
loess](https://rafalab.github.io/dsbook/smoothing.html#local-weighted-regression-loess)
for how kernel smoothers and loess work.  It is worth taking a look at his
animated graphics that show the repeated estimation across the length of the
data.

## Slide 18:  4 example plots with different kernel smoothers

<!--
# examples below copied from rafalab and edited
-->


```{r bin and kernel smoothing}
#---- bin and kernel smoothing ----

# kernel smoothing, box and normal kernel
# note the Gaussian is a smoother function, even for the same span
span <- 10 
fit_ru <- with(ts_ex_data, 
            ksmooth(time, y, kernel = "box", bandwidth = span))
fit_rw <- with(ts_ex_data, 
            ksmooth(time, y, kernel = "normal", bandwidth = span))
span <- 50
fit_su <- with(ts_ex_data, 
            ksmooth(time, y, kernel = "box", bandwidth = span))
fit_sw <- with(ts_ex_data, 
            ksmooth(time, y, kernel = "normal", bandwidth = span))

# 4 separate plots for now:
# Rough, unweighted
ts_ex_data %>% mutate(smooth = fit_ru$y) %>%
  ggplot(aes(time, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(time, smooth), color="red") +
  labs(title = "ARMA(1,1) time series\n box smooth (span 10)",
       x = "Time axis",
       y = "Outcome")

# Rough, weighted
ts_ex_data %>% mutate(smooth = fit_rw$y) %>%
  ggplot(aes(time, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(time, smooth), color="red")+
  labs(title = "ARMA(1,1) time series\n Gaussian smooth (span 10)",
       x = "Time axis",
       y = "Outcome")

# Smooth, unweighted
ts_ex_data %>% mutate(smooth = fit_su$y) %>%
  ggplot(aes(time, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(time, smooth), color="red") +
  labs(title = "ARMA(1,1) time series\n box smooth (span 50)",
       x = "Time axis",
       y = "Outcome")

# Smooth, weighted
ts_ex_data %>% mutate(smooth = fit_sw$y) %>%
  ggplot(aes(time, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(time, smooth), color="red") +
  labs(title = "ARMA(1,1) time series\n Gaussian smooth (span 50)",
       x = "Time axis",
       y = "Outcome")

```

## Slide 20:  loess

Loess is a locally weighted regression model fit over a small span. The default
in loewss is a 2nd degree polynomial, rather than a linear regression line.

<!--
# example code copied from rafalab link above and edited
-->
  
```{r loess example}
#---- loess example ----

# first convert to a tibble
ts_ex_data <- ts_ex_data %>% tibble()

# choose a span based on the length of the data
# default span is .75; we are using .1 (50/500)
span <- .25

# consider two different loess smoothers, linear and polynomial
fit1 <- loess(y ~ time, degree=1, span = span, data=ts_ex_data)
# the default is degree = 2
fit2 <- loess(y ~ time, span = span, data=ts_ex_data)

ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      smooth2 = fit2$fitted) %>%
  ggplot(aes(time, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(time, smooth1), color="red", lty = 1) +
  geom_line(aes(time, smooth2), color="blue", lty = 2) +
  labs(title = "ARMA(1,1) time series\n with loess smooth (span .25)",
       x = "Time axis",
       y = "Outcome")

# short span
span <- .1

# consider two different loess smoothers, linear and polynomial
fit1 <- loess(y ~ time, degree=1, span = span, data=ts_ex_data)
# the default is degree = 2
fit2 <- loess(y ~ time, span = span, data=ts_ex_data)

ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      smooth2 = fit2$fitted) %>%
  ggplot(aes(time, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(time, smooth1), color="red", lty = 1) +
  geom_line(aes(time, smooth2), color="blue", lty = 2) +
  labs(title = "ARMA(1,1) time series\n with loess smooth (span .1)",
       x = "Time axis",
       y = "Outcome")

# long span
span <- .75

# consider two different loess smoothers, linear and polynomial
fit1 <- loess(y ~ time, degree=1, span = span, data=ts_ex_data)
# the default is degree = 2
fit2 <- loess(y ~ time, span = span, data=ts_ex_data)

ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      smooth2 = fit2$fitted) %>%
  ggplot(aes(time, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(time, smooth1), color="red", lty = 1) +
  geom_line(aes(time, smooth2), color="blue", lty = 2) +
  labs(title = "ARMA(1,1) time series\n with loess smooth (span .75)",
       x = "Time axis",
       y = "Outcome")

```

# Slide 21:  Smooth vs. rough decomposition

```{r smooth vs. rough}
#---- smooth vs. rough ----

# after minimal smoothing -- long span
span <- .75

fit1 <- loess(y ~ time, degree=1, span = span, data=ts_ex_data)
ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      resid1  = y - fit1$fitted) %>%
  ggplot(aes(time, resid1)) +
  geom_point(size = .75, alpha = .5, color = "black") +
  geom_abline(intercept=0, slope = 0, color="red", lty = 1) +
  labs(title = "ARMA(1,1) residual time series\n light smoothing removed",
       x = "Time axis",
       y = "Residual Outcome")

# after more aggressive smoothing -- wigglier curve
span <- .1

fit1 <- loess(y ~ time, degree=1, span = span, data=ts_ex_data)

ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      resid1  = y - fit1$fitted) %>%
  ggplot(aes(time, resid1)) +
  geom_point(size = .75, alpha = .5, color = "black") +
  geom_abline(intercept=0, slope = 0, color="red", lty = 1) +
  labs(title = "ARMA(1,1) residual time series\n wiggly smoother removed",
       x = "Time axis",
       y = "Residual Outcome")


```

# Slide 22:  Differencing

```{r Differencing}
#---- Differencing ----

# first difference:
ts_ex_diff <- ts_ex_data %>% 
    mutate(diff = y - lag(y, default = first(y), order_by = time))

#ts_ex_diff

# plot the first differences and check how smooth they are:
span = 0.75
fit1 <- loess(y ~ time, degree=1, span = span, data=ts_ex_data)
ts_ex_diff %>% mutate(smooth1 = fit1$fitted,
                      resid1  = y - fit1$fitted) %>%
  ggplot(aes(time, diff)) +
  geom_point(size = .75, alpha = .5, color = "black") +
  geom_line(aes(time, smooth1), color="red", lty = 1) +
  labs(title = "ARMA(1,1) first differences\n w/ loess smooth (span .75)",
       x = "Time axis",
       y = "Residual Outcome")

```

# Spline smoothing

TODO:  Use this section to discus the difference between smoothing and regression splines and show some example smooths.
