---
title: "Week 9 Lecture Commands"
author: "Lianne Sheppard for ENVH 556"
date: "Winter 2021; Updated `r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
        fig_caption: yes
        toc: true
        toc_depth: 3
        number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#-----setup-----

# set knitr options
knitr::opts_chunk$set(echo = TRUE)

# clear work space of all objects and unload all extra (non-base) packages
rm(list = ls(all = TRUE))
if (!is.null(sessionInfo()$otherPkgs)) {
    res <- suppressWarnings(
        lapply(paste('package:', names(sessionInfo()$otherPkgs), sep=""),
               detach, character.only=TRUE, unload=TRUE, force=TRUE))
}

```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#-----load libraries pacman-----

# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.
pacman::p_load(knitr, dplyr, feasts, slider, tidyr, ggplot2)
# may not need older time series packages:
pacman::p_load(tseries, xts) 

```

# Slides 8 & 9:  AR(1) process example, $\alpha$=0.5

```{r AR(1) example}
#---- AR(1) example----

# n is length of time series
n <- 500
# generate data with alpha = .5
e <- rnorm(n)
y <- double(n)
y[1] <- 0
for (i in 2:n) {y[i] <- 0.5*y[i-1] + e[i]}
 
# make a tsibble 
ts_ex_data <- tibble(y, ind = seq(1,n)) %>%
  as_tsibble(index = ind)

# plot time series
ggplot() + 
  geom_line(data = ts_ex_data, aes(x = ind, y = y)) +
  labs(title = "AR(1) time series",
       x = "Time axis",
       y = "Outcome")
 
# plot autocorrelation
ts_ex_data %>% ACF(y) %>% autoplot()

# plot partial autocorrelation
ts_ex_data %>% PACF(y) %>% autoplot()

```

# Slide10:  AR(1) process example, $\alpha$=0.9

```{r AR(1) example2}
#---- AR(1) example2----

# n is length of time series
n <- 500
# generate data with alpha = .9
e <- rnorm(n)
y <- double(n)
y[1] <- 0
for (i in 2:n) {y[i] <- 0.9*y[i-1] + e[i]}
 
# make a tsibble 
ts_ex_data <- tibble(y, ind = seq(1,n)) %>%
  as_tsibble(index = ind)

# plot time series
ggplot() + 
  geom_line(data = ts_ex_data, aes(x = ind, y = y)) +
  labs(title = "AR(1) time series, alpha=0.9",
       x = "Time axis",
       y = "Outcome")
 
# plot autocorrelation
ts_ex_data %>% ACF(y) %>% autoplot()

# plot partial autocorrelation
ts_ex_data %>% PACF(y) %>% autoplot()

```

# Slide 13:  Moving average example

```{r MA(2) process example}
#---- MA(2) process example----

# n is length of time series
n <- 500

# generate data with rho1 = .0, rho2 = .8
e <- rnorm(n)
y <- double(n)
y[1] <- e[1]
y[2] <- e[2] + 0.8*e[1]
for (i in 3:n) {y[i] <- e[i] + 0.8*e[i-1] + 0.8*e[i-2]}
 
# make a tsibble 
ts_ex_data <- tibble(y, ind = seq(1,n)) %>%
  as_tsibble(index = ind)

# plot time series
ggplot() + 
  geom_line(data = ts_ex_data, aes(x = ind, y = y)) +
  labs(title = "MA(2) time series",
       x = "Time axis",
       y = "Outcome")
 
# plot autocorrelation
ts_ex_data %>% ACF(y) %>% autoplot()

# plot partial autocorrelation
ts_ex_data %>% PACF(y) %>% autoplot()



```

# Slide 14:  ARMA(1,1)

```{r ARMA(1,1)}
#---- ARMA(1,1) ----

set.seed(200)

# n is length of time series
n <- 500

# generate data with alpha=0.9, rho = .8
e <- rnorm(n)
y <- double(n)
y[1] <- e[1]
for (i in 2:n) {y[i] <- 0.9*y[i-1] + e[i] + 0.8*e[i-1] }
 
# make a tsibble 
ts_ex_data <- tibble(y, ind = seq(1,n)) %>%
  as_tsibble(index = ind)

# plot time series
ggplot() + 
  geom_line(data = ts_ex_data, aes(x = ind, y = y)) +
  labs(title = "ARMA(1,1) time series",
       x = "Time axis",
       y = "Outcome")
 
# plot autocorrelation
ts_ex_data %>% ACF(y) %>% autoplot()

# plot partial autocorrelation
ts_ex_data %>% PACF(y) %>% autoplot()

```

# Time plots with smoother overlaid

The next chunk attempts some ideas but so far none are working.  Would like to
show a running 10th percentile.

```{r Time plot with smoother, eval = FALSE}
#---- Time plot with smoother ----

# Chris: none of the following are working. Due to the tsibble? Maybe drop in
# favor of the kernel and loess smoothers below?  Thought it owuld be nice to
# get an example of the quantile smoother to work to accompany Brantley

# # Not working
# ts_ex_data <- ts_ex_data %>% 
#   #as_tibble() %>%
#   slide_dbl(ts_ex_data$y, ~mean(.x, na.rm = TRUE), .before = 10, .after = 10) 
#Example code:
#slide_dbl(x, ~mean(.x, na.rm = TRUE), .before = 1)

# # Not working:  fails at side %in% and says object 'side' not found
# # from here:  https://stackoverflow.com/questions/743812/calculating-moving-average
# moving_fn <- function(x, w, fun, ...) {
#   # x = vector with numeric data
#   # w = window length
#   # fun = function to apply
#   # side = side to take, (c)entre, (l)eft or (r)ight
#   # ... = parameters passed on to 'fun'
#   y <- numeric(length(x))
#   for (i in seq_len(length(x))) {
#     if (side %in% c("c", "centre", "center")) {
#       ind <- c((i - floor(w / 2)):(i + floor(w / 2)))
#     } else if (side %in% c("l", "left")) {
#       ind <- c((i - floor(w) + 1):i)
#     } else if (side %in% c("r", "right")) {
#       ind <- c(i:(i + floor(w) - 1))
#     } else {
#       stop("'side' must be one of 'centre', 'left', 'right'", call. = FALSE)
#     }
#     ind <- ind[ind %in% seq_len(length(x))]
#     y[i] <- fun(x[ind], ...)
#   }
#   y
# }
# 
# # and now any variation you can think of!
# moving_average <- function(x, w = 5, side = "centre", na.rm = FALSE) {
#   moving_fn(x = x, w = w, fun = mean, side = side, na.rm = na.rm)
# }
# 
# moving_sum <- function(x, w = 5, side = "centre", na.rm = FALSE) {
#   moving_fn(x = x, w = w, fun = sum, side = side, na.rm = na.rm)
# }
# 
# moving_maximum <- function(x, w = 5, side = "centre", na.rm = FALSE) {
#   moving_fn(x = x, w = w, fun = max, side = side, na.rm = na.rm)
# }
# 
# moving_median <- function(x, w = 5, side = "centre", na.rm = FALSE) {
#   moving_fn(x = x, w = w, fun = median, side = side, na.rm = na.rm)
# }
# 
# moving_Q1 <- function(x, w = 5, side = "centre", na.rm = FALSE) {
#   moving_fn(x = x, w = w, fun = quantile, side = side, na.rm = na.rm, 0.25)
# }
# 
# moving_Q3 <- function(x, w = 5, side = "centre", na.rm = FALSE) {
#   moving_fn(x = x, w = w, fun = quantile, side = side, na.rm = na.rm, 0.75)
# }
# 
# moving_10th <- function(x, w = 5, side = "centre", na.rm = FALSE) {
#   moving_fn(x = x, w = w, fun = quantile, side = side, na.rm = na.rm, 0.10)
# }

# # Test rolling background when the above finally works:
# roll10_y <- moving_10th(ts_ex_data$y, w = 10, side = "c")

```

# Loess and kernel smoother examples

Examples in this section follow the nice exposition by Rafael A. Irizarry in his
[Introduction to Data Science](https://rafalab.github.io/dsbook/) on [bin
smoothing](https://rafalab.github.io/dsbook/smoothing.html#bin-smoothing) and [
loess](https://rafalab.github.io/dsbook/smoothing.html#local-weighted-regression-loess)
for how kernel smoothers and loess work.  It is worth taking a look at his
animated graphics that show the repeated estimation across the length of the
data.

## Slide 18:  4 example plots with different kernel smoothers

<!--
# examples copied from rafalab:

# # bin smoothing
# span <- 7 
# fit <- with(polls_2008, 
#             ksmooth(day, margin, kernel = "box", bandwidth = span))
# 
# polls_2008 %>% mutate(smooth = fit$y) %>%
#   ggplot(aes(day, margin)) +
#     geom_point(size = 3, alpha = .5, color = "grey") + 
#   geom_line(aes(day, smooth), color="red")

# # kernel smoothing, normal kernel
# span <- 20
# fit <- with(polls_2008, 
#             ksmooth(day, margin, kernel = "normal", bandwidth = span))
# 
# polls_2008 %>% mutate(smooth = fit$y) %>%
#   ggplot(aes(day, margin)) +
#   geom_point(size = 3, alpha = .5, color = "grey") + 
#   geom_line(aes(day, smooth), color="red")
-->


```{r bin and kernel smoothing}
#---- bin and kernel smoothing ----


# # TODO: Optional; what is wrong with this?  The separate code
# # (sample below) works but this faceted version doesn't
# # examples implemented:
# # bin smoothing
# span <- 10 
# fit_ru <- with(ts_ex_data, 
#             ksmooth(ind, y, kernel = "box", bandwidth = span))
# fit_rw <- with(ts_ex_data, 
#             ksmooth(ind, y, kernel = "normal", bandwidth = span))
# span <- 50
# fit_su <- with(ts_ex_data, 
#             ksmooth(ind, y, kernel = "box", bandwidth = span))
# fit_sw <- with(ts_ex_data, 
#             ksmooth(ind, y, kernel = "normal", bandwidth = span))
# 
# data_smooth <- ts_ex_data %>% tibble() %>%
#   # create new variables from the different fits
#   mutate(smooth_ru = fit_ru$y,
#          smooth_rw = fit_rw$y,
#          smooth_su = fit_su$y,
#          smooth_sw = fit_su$y) %>%
#   # make dataset longer
#   pivot_longer(cols = c(contains("smooth_")), 
#                names_to = "Smoother_type", 
#                values_to = "smooth_fit")
# 
#   ggplot(aes(ind, y=smooth_fit)) +
#     # different plots by error type
#     facet_wrap(~error_type) +
#     # raw data as points
#     geom_point(size = 1, alpha = .5, shape = "o", color = "grey") + 
#     # lines for smooths
#     geom_line(aes(ind, smooth_fit), color="red")
# 

# kernel smoothing, box and normal kernel
# note the Gaussian is a smoother function, even for the same span
span <- 10 
fit_ru <- with(ts_ex_data, 
            ksmooth(ind, y, kernel = "box", bandwidth = span))
fit_rw <- with(ts_ex_data, 
            ksmooth(ind, y, kernel = "normal", bandwidth = span))
span <- 50
fit_su <- with(ts_ex_data, 
            ksmooth(ind, y, kernel = "box", bandwidth = span))
fit_sw <- with(ts_ex_data, 
            ksmooth(ind, y, kernel = "normal", bandwidth = span))

# 4 separate plots for now:
# Rough, unweighted
ts_ex_data %>% mutate(smooth = fit_ru$y) %>%
  ggplot(aes(ind, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(ind, smooth), color="red") +
  labs(title = "ARMA(1,1) time series\n box smooth (span 10)",
       x = "Time axis",
       y = "Outcome")

# Rough, weighted
ts_ex_data %>% mutate(smooth = fit_rw$y) %>%
  ggplot(aes(ind, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(ind, smooth), color="red")+
  labs(title = "ARMA(1,1) time series\n Gaussian smooth (span 10)",
       x = "Time axis",
       y = "Outcome")

# Smooth, unweighted
ts_ex_data %>% mutate(smooth = fit_su$y) %>%
  ggplot(aes(ind, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(ind, smooth), color="red") +
  labs(title = "ARMA(1,1) time series\n box smooth (span 50)",
       x = "Time axis",
       y = "Outcome")

# Smooth, weighted
ts_ex_data %>% mutate(smooth = fit_sw$y) %>%
  ggplot(aes(ind, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(ind, smooth), color="red") +
  labs(title = "ARMA(1,1) time series\n Gaussian smooth (span 50)",
       x = "Time axis",
       y = "Outcome")


```

## Slide 20:  loess

Loess is a locally weighted regression model fit over a small span. The default
in loewss is a 2nd degree polynomial, rather than a linear regression line.

<!--
# example code copied from rafalab link above
total_days <- diff(range(polls_2008$day))
span <- 21/total_days

fit <- loess(margin ~ day, degree=1, span = span, data=polls_2008)

polls_2008 %>% mutate(smooth = fit$fitted) %>%
  ggplot(aes(day, margin)) +
  geom_point(size = 3, alpha = .5, color = "grey") +
  geom_line(aes(day, smooth), color="red")
  -->
  
```{r loess example}
#---- loess example ----

# first convert to a tibble
ts_ex_data <- ts_ex_data %>% tibble()

# choose a span based on the length of the data
# default span is .75; we are using .1 (50/500)
span <- .25

# consider two different loess smoothers, linear and polynomial
fit1 <- loess(y ~ ind, degree=1, span = span, data=ts_ex_data)
# the default is degree = 2
fit2 <- loess(y ~ ind, span = span, data=ts_ex_data)

ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      smooth2 = fit2$fitted) %>%
  ggplot(aes(ind, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(ind, smooth1), color="red", lty = 1) +
  geom_line(aes(ind, smooth2), color="blue", lty = 2) +
  labs(title = "ARMA(1,1) time series\n with loess smooth (span .25)",
       x = "Time axis",
       y = "Outcome")

# short span
span <- .1

# consider two different loess smoothers, linear and polynomial
fit1 <- loess(y ~ ind, degree=1, span = span, data=ts_ex_data)
# the default is degree = 2
fit2 <- loess(y ~ ind, span = span, data=ts_ex_data)

ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      smooth2 = fit2$fitted) %>%
  ggplot(aes(ind, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(ind, smooth1), color="red", lty = 1) +
  geom_line(aes(ind, smooth2), color="blue", lty = 2) +
  labs(title = "ARMA(1,1) time series\n with loess smooth (span .1)",
       x = "Time axis",
       y = "Outcome")

# long span
span <- .75

# consider two different loess smoothers, linear and polynomial
fit1 <- loess(y ~ ind, degree=1, span = span, data=ts_ex_data)
# the default is degree = 2
fit2 <- loess(y ~ ind, span = span, data=ts_ex_data)

ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      smooth2 = fit2$fitted) %>%
  ggplot(aes(ind, y)) +
  geom_point(size = 1, alpha = .5, color = "grey") +
  geom_line(aes(ind, smooth1), color="red", lty = 1) +
  geom_line(aes(ind, smooth2), color="blue", lty = 2) +
  labs(title = "ARMA(1,1) time series\n with loess smooth (span .75)",
       x = "Time axis",
       y = "Outcome")

```

# Slide 21:  Smooth vs. rough decomposition

```{r}
#---- smooth vs. rough ----

# after minimal smoothing -- long span
span <- .75

fit1 <- loess(y ~ ind, degree=1, span = span, data=ts_ex_data)
ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      resid1  = y - fit1$fitted) %>%
  ggplot(aes(ind, resid1)) +
  geom_point(size = .75, alpha = .5, color = "black") +
  geom_abline(intercept=0, slope = 0, color="red", lty = 1) +
  labs(title = "ARMA(1,1) residual time series\n light smoothing removed",
       x = "Time axis",
       y = "Residual Outcome")

# after more aggressive smoothing -- wigglier curve
span <- .1

fit1 <- loess(y ~ ind, degree=1, span = span, data=ts_ex_data)

ts_ex_data %>% mutate(smooth1 = fit1$fitted,
                      resid1  = y - fit1$fitted) %>%
  ggplot(aes(ind, resid1)) +
  geom_point(size = .75, alpha = .5, color = "black") +
  geom_abline(intercept=0, slope = 0, color="red", lty = 1) +
  labs(title = "ARMA(1,1) residual time series\n wiggly smoother removed",
       x = "Time axis",
       y = "Residual Outcome")


```

# Slide 22:  Differencing

```{r}
#---- Differencing ----

# first difference:
ts_ex_diff <- ts_ex_data %>% 
    mutate(diff = y - lag(y, default = first(y), order_by = ind))

#ts_ex_diff

# plot the first differences and check how smooth they are:
span = 0.75
fit1 <- loess(y ~ ind, degree=1, span = span, data=ts_ex_data)
ts_ex_diff %>% mutate(smooth1 = fit1$fitted,
                      resid1  = y - fit1$fitted) %>%
  ggplot(aes(ind, diff)) +
  geom_point(size = .75, alpha = .5, color = "black") +
  geom_line(aes(ind, smooth1), color="red", lty = 1) +
  labs(title = "ARMA(1,1) first differences\n w/ loess smooth (span .75)",
       x = "Time axis",
       y = "Residual Outcome")

```


