---
title: "Week3Lecture"
author: "Lianne Sheppard"
date: "Revised for Winter 2021; Updated `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

<!--Basic document set-up goes here  -->
```{r setup, include=FALSE}
#-----setup----
knitr::opts_chunk$set(echo = TRUE)
```

```{r load.libraries.pacman, echo=FALSE, include=FALSE, eval=TRUE}
#-----load.libraries.pacman------------
# Load pacman into memory, installing as needed
my_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_repo)}

# Load the other packages, installing as needed.  Some reasons for packages:
# knitr:  kable()
# tidyverse: multiple data science packages including ggplot, dplyr, readr, broom, tidyr, purr -- only loading those we expect to use
# multcomp:  glht
# car:  linearHypothesis
pacman::p_load(dplyr, ggplot2, broom, tidyr, purr, knitr, multcomp, dplyr, modelr, car)  
```

```{r read.data, echo=FALSE, warning=FALSE}
#-----read data--------

# get working directory
project_path <- getwd()

# create data directory, if needed    
dir.create(file.path(project_path,"Datasets"), showWarnings=FALSE, recursive = TRUE) 

# create the path for the Datasets directory
data_path <- file.path(project_path,"Datasets")

# specify file name and path
file_name <- "allseasonsR.rds"
file_path <- file.path(data_path, file_name)

# Only download the file if it's not already present
if (!file.exists(file_path)) {
    url <- paste("https://staff.washington.edu/high/envh556/Datasets", 
                 file_name, sep = '/')
    download.file(url = url, destfile = file_path)
}

# Output a warning message if the file cannot be found
if (file.exists(file_path)) {
    snapshot <- readRDS(file = file_path)
} else warning(paste("Can't find", file_name, "!"))

# make dataframe a tibble
snapshot <- as_tibble(snapshot)

# remove temporary variables
rm(url, file_name, file_path)

```


This is a companion document to the Week 3 lecture.  Each slide with R code is represented in this document.  This document also has a few additional alternative coding options.

## Slide 10:  Simple regression scatterplot ln_nox on ln_no

```{r slide10}
#-----slide10----

(p <- ggplot(data=snapshot,aes(ln_no,ln_nox)) +
     geom_point() +
     labs(
        title="Association between NO and NOx on the log scale",
        x="ln(NO) (ln(ppb))",
        y="ln(NOx) (ln(ppb))"
    ) )

```


## Slide 12:  Add regression line

```{r slide12}
#-----slide12----

p  + stat_smooth(method=lm, se=FALSE)

```

## Slide 13 + 14:  Linear regression model

```{r slide13+14}
#-----slide13+14----

# fit model
nox.fit <- lm(ln_nox ~ ln_no, data = snapshot)

# print summary, coefficients and CIs
summary(nox.fit)
coef(nox.fit)
confint(nox.fit)

# variance-covariance matrix
vcov(nox.fit)

# # "tidy" option for summary, coefficients, and CIs
# from the broom package
summary_tbl <- tidy(nox.fit, conf.int = TRUE, conf.level = 0.95)
summary_tbl

```

## Slide 18

```{r slide18}
#-----slide18----

# Should be basic description of the factor variable.
# season is the same variable as seasonfac, not as a factor and coded the same
# 1,2,3. Verify this by looking at the data:
summary(snapshot$seasonfac)
summary(snapshot$season)
table(snapshot$season)
table(snapshot$seasonfac)
table(snapshot$seasonfac,snapshot$season)

```

## Slide 19

```{r slide19}
#-----slide19----

nox.aov<-lm(ln_nox ~ seasonfac, data = snapshot)
anova(nox.aov)
summary(nox.aov)  

#model.matrix(nox.aov)
margin.table(model.matrix(nox.aov), 2)

```

## Slide 20:  add means of ln_nox by season 

For example of interpreting dummy variables

```{r slide20}
#-----slide20----

# Simple row of means of ln_nox by season
round(tapply(snapshot$ln_nox, snapshot$seasonfac, mean), 2)

# # dplyr option (same results, output is formatted differently)
snapshot %>% 
    group_by(seasonfac) %>% 
    summarise(mean(ln_nox), .groups = "drop") %>% 
    mutate_if(is.numeric, round, 2)

# full table of N, means & sd of ln_nox by season, dplyr option
snapshot %>%
    group_by(seasonfac) %>%
    summarise(N = n(), 
              mean = mean(ln_nox),
              sd = sd(ln_nox), 
              .groups = "drop") %>%
    mutate_if(is.double, round, 2)

```


## Slide 21:  Be aware of coding!

```{r slide21}
#-----slide21----

nox.aov2<-lm(ln_nox~season, data=snapshot)
anova(nox.aov2)
summary(nox.aov2)  

```

## Slide 22:  Season-stratified analysis example:  means of ln_nox, ln_no by season


```{r define.summary.function}
#-----define summary function----

summary_stats <- function(df = snapshot, summary_var, grouping_var = seasonfac, 
                          digits = 2){
  
  # specify names & variables for non-standard evaluation
  grouping_var <- enquo(grouping_var)
  summary_var <- enquo(summary_var)
  summary_name <- quo_name(summary_var)
  
  # create names for summary stats
  summary_name_N <- paste0(summary_name, ".N")
  summary_name_mean <- paste0(summary_name, ".mean")
  summary_name_sd <- paste0(summary_name, ".sd")
  
  # create summary
  df %>%
    group_by(!!grouping_var) %>% 
    summarise(!!summary_name_N := n(), 
              !!summary_name_mean := mean(!!summary_var), 
              !!summary_name_sd := sd(!!summary_var), 
              .groups = "drop") %>% 
    mutate_if(is.double, round, digits)
}

```


```{r slide 22}
#-----slide22-------------------------

# ln_nox
summary_stats(summary_var = ln_nox)

# ln_no
summary_stats(summary_var = ln_no)

# ln_nox and ln_no
full_join(summary_stats(summary_var = ln_nox), 
          summary_stats(summary_var = ln_no), 
          by = "seasonfac")

```


## Slide 23:  Scatterplot with 3 season-specific lines

```{r slide23, message=FALSE}
#-----slide23----

# ggplot with separate colors and lines by season
ggplot(snapshot, mapping = aes(x = ln_no, y = ln_nox, color = seasonfac)) +
    geom_point(alpha = .4, size=.7) +
    stat_smooth(method = lm, se=FALSE) +
    #geom_rug(position="jitter", size=.2) +
    xlim(range(snapshot$ln_no, snapshot$ln_nox)) +
    ylim(range(snapshot$ln_no, snapshot$ln_nox)) +
    geom_abline(slope = 1, intercept = 0, alpha = .2, color = "black") +
    coord_fixed() +
    labs(x="ln(NO) (ln(ppb))", y="ln(NOx) (ln(ppb)", color="Season") + 
    theme_bw()

```


## Slide 25:  Season-specific regression

Summarize each season-specific model

```{r slide25}
#-----slide25----

# season-specific regressions, as shown on slide (only subset of output)
summary(lm(ln_nox ~ ln_no, data = snapshot, subset = seasonfac == "1Summer"))
summary(lm(ln_nox ~ ln_no, data = snapshot, subset = seasonfac == "2Fall"))
summary(lm(ln_nox ~ ln_no, data = snapshot, subset = seasonfac == "3Winter"))

```

```{r slide25.dplyr, eval = FALSE}
#----slide25.dplyr----

# # dplyr option -- CHRIS: currently not working
# Need to load the broom, purr, & tidyr packages for this approach
season_mdls <- snapshot %>% 
  
  # group by season
  group_by(seasonfac) %>%
  
  # nest the tibble
  nest() %>% 
  
  # run regression model for each group, clean up with `tidy()`
  mutate(model = map(data, ~lm(ln_nox ~ ln_no, data = .x)),
         summary = map(model, summary),
         tidied = map(model, tidy) 
         ) %>% 
  
  # unnest
  unnest(tidied)
 
# show seasonal model summaries
season_mdls$summary

```


## Slide 26:  Season-adjusted regression

```{r slide26.season.adjusted}
#-----slide26 season adjusted----

# season-adjusted regression
nox.fit.seasadj <-lm(ln_nox ~ ln_no + seasonfac, data = snapshot)

summary(nox.fit.seasadj)
confint(nox.fit.seasadj)

```


## Slide 27:  Be aware of coding! 

Season as a continuous variable


```{r slide27}
#-----slide27----

# Season as a continuous variable

# Fit the linear regression with continuous season variable
summary(lm(ln_nox ~ ln_no + season, data = snapshot))

# Compare to the regression on the previous slide
 
```


## Slide 29:  Interaction model

Need to load the `car` package for the `linearHypothesis` command.  It masks at least one object from several other packages, so don't load if you don't need it.  You can explicitly tell R which package to use with the "packagename::" prefix to a function or object.

```{r slide29}
#-----slide29----

# interaction model
snap_int <- lm(ln_nox ~ ln_no * seasonfac, data = snapshot)
summary(snap_int)

# F test for significant interaction from the car package
linearHypothesis(snap_int, c("ln_no:seasonfac2Fall=0", "ln_no:seasonfac3Winter=0"), 
                 test="F")

```


## Slide 30:  "lincom" coeficient estimates

Note:  This can be done using the `glht` command from the `multcomp` package.  This will mask several objects from other packages including `select` from `dplyr`, and `boxcox` from `EnvStats`.

Here is an example I found online: 
```
# mod is the output of a lm
# Extract the coefficient names (b0,b1,b2,b3)
names(coef(mod))  
# linfct specifies the required combination In this case we want b1+b3=0
# (effect of gender in age>40)
summary(glht(mod, linfct = c("genderMale + genderMale:agehiTRUE = 0")))
# Show the confidence interval
mod.lh <- glht(mod, linfct = c("genderMale + genderMale:agehiTRUE = 0"))
confint(mod.lh)

```

```{r find.means}
#-----find means----

# Note:  this does not require snapshot to be a tibble
snapshot %>%
    group_by(seasonfac) %>%
    summarise(N = length(ln_nox), 
              ln_no_mean = mean(ln_no),
              ln_nox_mean = mean(ln_nox),
              ln_nox_sd = sd(ln_nox),
              ln_nox_LCI = ln_nox_mean - 1.96*ln_nox_sd / sqrt(N),
              ln_nox_UCI = ln_nox_mean + 1.96*ln_nox_sd / sqrt(N) 
              )

```


```{r slide30.lincom.example}
#-----slide30 lincom example----

# Estimate the mean ln(NOx) when ln(NO) is 4

# Show the coefficient names.  Their order is needed for the K matrix definition
# so we know which values go in which column
names(coef(snap_int))

# K is the matrix of contrasts of interest to create the linear combination of
# the coefficients times the values in K.  Each row is for a contrast of
# interest.  The values in the rows are ordered by the coefficients. First row
# is summer when ln_no is 4, then same question for fall, then winter
K <- rbind(summer = c(1,4,0,0,0,0),
           fall   = c(1,4,1,0,4,0),
           winter = c(1,4,0,1,0,4))

colnames(K) <- names(coef(snap_int))

K

# estimate the summer, fall, winter effects simultaneously
season_effects <- glht(snap_int, linfct = K)
summary(season_effects)

# and get its CI
confint(season_effects)

```

## Slide 32:  example residual analysis plots

```{r slide32.default.residuals}
#-----Slide32 default residuals----

# IDRE example from https://stats.idre.ucla.edu/r/dae/robust-regression/
opar <- par(mfrow = c(2,2), oma = c(0, 0, 1.1, 0))
plot(snap_int, las = 1)

```

```{r a.basic.residual.plot, message=FALSE}
#-----a basic residual plot----

# first add data to the snapshot data frame with `modlr` functions
snapshot2 <- snapshot %>%
    add_residuals(snap_int,"resids") %>%
    add_predictions(snap_int,"preds")

# then make scatterplot
ggplot(snapshot2, aes(x = preds, y = resids)) +
    geom_point() +
    geom_abline(slope = 0, intercept = 0) +
    geom_smooth(method = "loess")

```


## Slide 35:  Regression for prediction -- starting to replicate Table 4 of Mercer et al Summer

```{r slide35}
#-----slide35----  

# replicate the summer model in Table 4
# First create a summer only dataset with only the variables of interest

# specify covariates of interest (omitting outcome variable)
vars <- c("D2A1", "A1_50", "A23_400", "Pop_5000", "D2C", "Int_3000", "D2Comm")

# create dataframe
summer <- snapshot %>%
    dplyr::filter(season == 1) %>%
    dplyr::select(ln_nox, all_of(vars))

# specify regression formula
frml <- as.formula(paste("ln_nox ~ ", paste(vars, collapse = "+") ) )

# fit regression model
summer_mod <- lm(frml, data = summer)

# show the results
summary(summer_mod)

```



## Slide 36:  Regression for prediction:  getting fitted values and SE of mean vs new value PLUS show R2=square of correlation

Resulting correlation should be about .85 to get the estimated R2 of .7252.  

```{r slide36}
#-----slide36 fitted values and residuals----

#Note:  Dropping the use of a new "summer2" dataframe as not needed.  Slides introduce a new data frame

# Then add the predictions (and the residuals to the dataset)
summer <- summer %>%  
    add_residuals(summer_mod,"resids") %>%
    add_predictions(summer_mod,"preds")

# and estimate the correlation
r <- cor(summer$preds, summer$ln_nox)
insamp_R2 <- r^2
paste("In-sample R2:  ", insamp_R2)
paste("correlation:  ", r)

```


## Slide 37:  Plot of regression with prediction interval, and add alongside the one with the CI

```{r slide37, message=FALSE}
#-----slide 37: plots w/ prediction intervals----

# in ggplot it is easy to get CIs of fitted lines but more work to get prediction intervals.  We show both here.
# Note:  add_predictions doesn't also add uncertainty of predictions
# first get the prediction intervals
# Note:  you can do this twice on the same dataset and get the same 4 variables
# appended twice.  If you use different interval options (interval="confidence"
# vs. interval="prediction" you will get different values in the created
# variables "lwr" and "upr" corresponding to the type of interval.  Note: you
# can't use a dataframe with multiple variables with the same name.
 
# predict creates the variables fit, lwr, upr
summer2 <- cbind(summer, predict(summer_mod, interval = "prediction"))

# get the plotting range
summer_mod_range <- with(summer2, range(ln_nox, lwr, upr))

# plot with 95% CI on the fitted line, uses default ggplot option
ggplot(summer2, aes(x = fit, y = ln_nox)) +
    geom_point() +
    stat_smooth(method = lm, se = TRUE) +
    coord_fixed() +
    xlim(summer_mod_range) + 
    ylim(summer_mod_range) +
    labs(title = "Mercer et al (2011) Table 4 Summer model fit:\npredictions vs. observations (ln(ppb)) and 95% CI", 
         x = "predicted ln(NOx) (ln(ppb))",
         y = "observed ln(NOx) (ln(ppb))"
         ) + 
    theme_bw()

# plot with 95% PI on the fitted line, need extra code to handle this 
ggplot(summer2, aes(fit)) +
    geom_point(aes(y = ln_nox)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "blue", alpha = 0.2) +
    geom_line(aes(y = fit), color = "blue", size = 1) +
    coord_fixed() +
    xlim(summer_mod_range) + 
    ylim(summer_mod_range) + 
    labs(title = "Mercer et al (2011) Table 4 Summer model fit:\npredictions vs. observations (ln(ppb)) and 95% PI",
         x = "predicted ln(NOx) (ln(ppb))",
         y = "observed ln(NOx) (ln(ppb))"
         ) + 
    theme_bw()

```

